---
title: "Assimilative causal inference (ACI) in R"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Assimilative causal inference (ACI) in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 3.6, collapse = TRUE,
                      comment = "#>")
library(aci)
set.seed(1)
```

## 1. Answering causal questions

### Estimating causal effects
Causal inference techniques provide useful tools for answering questions about
causality using observational data, including tasks that require estimating the
size of treatment effects. Many such frameworks and methods exist for causal effect
estimation, including graphical modelling approaches for estimating the causal effect
of a treatment (the cause) on an outcome (the effect). Such approaches require 
observable causes and their effect(s), captured in the data, to identify a causal estimand.


### Tracing backward from effects to causes
Assimilative causal inference (ACI) (Andreou et al., 2026) is a complementary
approach that formulates causality as an inverse problem, using Bayesian data 
assimilation to trace causes backward from their observed effects. This approach
can detect if one variable (a latent cause) influences another (an observed effect),
and if so, for how long a causal influence persists.

### The causal influence range (CIR)

Andreou & Chen (2026) develop the causal influence range (CIR) for quantifying 
the duration of a causal effect in both forward and backward directions. 
This answers questions such as: (1) "for how long will a causal effect persist?" 
(forward CIR); and (2) "when did the causal precursors of this event emerge?" 
(backward CIR). 

## 2. Getting started with ACI in R

### A different way of thinking about modelling
ACI requires a dynamical model to identify instantaneous causal relationships (Andreou et al., 2026).
In a standard regression workflow the model is the output: `lm()` is handed
data and returns coefficients. Here the model is the input: a statement of how
the observed and hidden variables evolve together, and how noisy each is,
supplied before the causal question is asked. Fitting still happens (in
Section 3, `aci_fit()` learns the model from data), but what it produces is
the object the causal question is answered relative to, not the answer itself.
A model can also be written down directly (`cgns_model()`) or calibrated to
data (`calibrate_cgns()`); the advanced vignette describes both routes.

### The question ACI answers
Some variables can be observable while others remain hidden, and for a latent
process $y$ (the cause) influencing observed variables 
(the effect) in a time series: **is a causal influence detected, and if so, how 
strongly at each moment, and for how long?**

To answer this question, we can compare two reconstructions of $y_t$:

* the **filter** $p(y_t \mid x_{\le t})$: what past observations say;
* the **smoother** $p(y_t \mid x_{\le T})$: what past *and future* observations say.

If the future of $x$ sharpens our knowledge of $y_t$, then it confirms the existence
of a causal influence. 

The **ACI metric** is the relative entropy between the two posteriors,

$$\mathrm{ACI}(t) \;=\; D_{\mathrm{KL}}\!\big(p^{\text{smoother}}_t \,\big\|\,
p^{\text{filter}}_t\big) \;\ge\; 0,$$

computed *along one trajectory* (causality as it happened, not on average).


### Simulating example data

Since the model comes first, we can simulate from it and hold the hidden
truth aside to check the reconstruction against.

```{r quickstart-data}
truth <- model_dyad(params = list(d_x = 1, gamma = 2, f_x = 1.5, s_x = 0.7,
                                  d_y = 1, f_y = 0.5, s_y = 2))
sim <- simulate(truth, seed = 8, T = 5, dt = 5e-3, burn_in = 1)
dat <- data.frame(t = sim$obs$t, x1 = sim$obs$x[, 1], y = sim$hidden[, 1])
head(dat, 3)
```

## 3. Fit a model

Supply a formula naming the latent cause and the observed effects to `aci_fit()`,
along with a data frame. 

`aci_fit()` learns a conditional Gaussian model and returns a single object.

```{r quickstart-fit}
fit <- aci_fit(y ~ x1, data = dat)
fit
```

## 4. Plot and print the results

`plot()` visualises the reconstruction and the metric:

```{r quickstart-latent}
plot(fit, "latent", truth = dat$y)
legend("topright", c("reconstruction ± 2 sd", "withheld truth"),
       col = c(1, 2), lty = 1, bty = "n")
```

```{r quickstart-metric}
plot(fit, "metric")
```

`summary()` shows the learned equations and, with `verdict = TRUE`, a
significance test against the decoupled null, the fitted model with its
coupling removed (see the significance and validation section of
`vignette("vignette-2-advanced")`):

```{r quickstart-summary}
suppressWarnings(summary(fit, verdict = TRUE, B = 9))
```

## 5. Obtain the CIR

The causal influence range (CIR) (*for how long does $y$ influence $x_1$?*) 
is obtained using `cir()`:

```{r quickstart-cir}
cir(fit)
plot(fit, "cir")
```

Each row is one anchor time `t`: `tau` is the range in the model's time units
and `strength` is the peak finite-lag divergence there. `cir(fit)` reports the
forward direction; the backward question of Section 1 is
`cir(fit, direction = "backward")`.

## 6. When the cause was never measured

Drop the `y` column and `aci_fit()` switches to partial-observation mode:
the latent path is reconstructed by data assimilation around a seed model,
then the dynamics are learned from posterior samples. This mode is
**experimental** (single learning pass; see `?learn_model` notes).

```{r partial, fig.height=3.2}
fitp <- aci_fit(y ~ x1, data = dat[, c("t", "x1")], n_samples = 8)
plot(fitp, "latent", truth = dat$y)
```

We're well on our way to seeing what ACI can do. The underlying machinery is
covered in `vignette("vignette-2-advanced")`.
