---
title: "The ACI engine: from models to extremes"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{The ACI engine: from models to extremes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 3.6, collapse = TRUE,
                      comment = "#>")
library(aci)
set.seed(1)
```

## 1. Models and observations

ACI uses a class of models called **conditional Gaussian nonlinear systems
(CGNS)**: given the observed path the hidden dynamics are linear-Gaussian,
which gives a closed-form filter and smoother.

The package includes benchmark models from the ACI literature; the full
catalogue is summarised in the README and on each constructor's help page.
The dyad is the running example:

```{r benchmark-models}
m <- model_dyad()   # andreou2026aci Sec 3.1 / SI.4.1 defaults, MATLAB-checked
m
```

`simulate()` integrates the SDE (Euler-Maruyama) and returns the observed
trajectory and the hidden truth separately, so reconstructions can be scored
against the truth:

```{r simulate}
sim <- simulate(m, seed = 7, T = 6, dt = 5e-3, burn_in = 1)
str(sim$obs, max.level = 1)
```

For your own data, `observed_trajectory(t, x)` builds the observation object
(uniform grid, noise-free observations of the resolved variables).

Build custom mechanistic models with `cgns_model()` (explicit coefficients)
or `cgns_from_affine()` for a full drift; the latter checks conditional
linearity and splits the drift accordingly. `calibrate_cgns()` fits a bounded
subset of a user-supplied builder's parameters by deterministic least
squares, then estimates observed-channel noise from increment residuals (a
package extension).


## 2. The engine: filter, smoother, metric

```{r engine}
init <- list(mean = 0, cov = diag(1, 1))
filt <- da_filter(m, sim$obs, init = init)
smoo <- da_smooth(m, sim$obs, filter = filt)
plot(smoo, truth = sim$hidden[, 1])
```

`aci()` combines the pair into the metric and names its two parts: a
**signal** term, from disagreement between the posterior means, and a
**dispersion** term, from disagreement between the posterior spreads:

```{r metric}
aci_out <- aci(m, sim$obs, init = init)
plot(aci_out)
```

Numerical note: the default explicit stepper is the reference scheme and the
convention the lag table in Section 3 assumes; the CIR machinery rejects
`stepper = "implicit"` outright, and discards a supplied implicit filter to
rebuild an explicit one with a warning (`cir()` on an `aci_fit` rebuilds
silently). For stiff learned models or very
diffuse priors, switch to the positivity-preserving implicit stepper:
`da_filter(..., stepper = "implicit", nsub = 5)`; `aci_fit()` takes the same
two arguments.


## 3. The causal influence range (CIR)

The ACI metric says *how strongly now*. The **causal influence range
(CIR)** says *for how long*: at each time $t$, how far into the future do
observations still carry information that $y_t$ planted?

### The forward CIR

`lag_table()` stores the finite-lag divergences plus a heuristic tail
estimate; cost is proportional to the retained lag cells, up to $O(N^2)$ for
a full table. `forward_cir()` reads off the exact ("layer-cake") length, or
the cheaper ratio bound that the ACI reference code calls its efficient
approximation. Both integrate with composite Simpson: the exact form always, and the
ratio by default, following the ACI and EnKBS codebases rather than the plain
L1 grid sum on the FBCIR scripts' active line. The table below is adaptively
truncated (`tol = 1e-4`); the bound labels record this in `fc$bound` as
`objective_on_truncated_table` and `lower_ratio_on_truncated_table_only`,
and `tol = 0` reproduces the
untruncated convention of the published figures:

```{r cir-forward}
tab <- lag_table(m, sim$obs, mode = "forward", tol = 1e-4, init = init)
fc  <- forward_cir(tab)
l1  <- forward_cir(tab, method = "l1_linf")
plot(fc$t, fc$tau, type = "l", xlab = "t", ylab = expression(tau[f]),
     ylim = c(0, 1.3 * max(fc$tau, na.rm = TRUE)),
     main = "Forward CIR: exact vs efficient approximation")
lines(l1$t, l1$tau, col = 4, lty = 2)
legend("top", c("exact", "l1/linf (underestimate)"), col = c(1, 4),
       lty = 1:2, bty = "n", horiz = TRUE)
```

The final anchor is masked: the forward metric is exactly zero at the final
time. The decline through the end of the record is a finite record artefact of
those short rows, since no anchor can report a range longer than the
observations that follow it; the reference code sidesteps this with a lookahead
margin. A lag table uses the complete online Theorem 3 smoother
(Andreou & Chen, 2026) as its reference rather than the backward ODE smoother
`da_smooth()` returns, so their finite-grid diagonals can differ at finite
$\Delta t$.

### The subjective CIR

The $\varepsilon$-resolved **subjective CIR** (how far ahead the influence
last exceeds each threshold) is reached through the `eps` argument
(Andreou & Chen, 2026):

```{r cir-subjective, fig.height=3.2}
eps <- 10^seq(-4, 0, length.out = 60)
sc  <- forward_cir(tab, eps = eps)
filled.contour(sc$t, log10(eps), sc$subjective, xlab = "t",
               ylab = expression(log[10]~epsilon),
               main = "Subjective CIR length",
               color.palette = function(n) hcl.colors(n, "viridis"))
```

### The backward CIR

The **backward** direction (*when did the causal precursors of this event
emerge?*) is the attribution estimand:

```{r cir-backward}
b <- backward_cir(m, obs = sim$obs, T = c(2.5, 4, 5.5), init = init)
data.frame(T = b$t, tau = b$tau, M = b$M,
           from = b$interval[, 1], to = b$interval[, 2])
```

For a reference anchor $T$, the one-observation discrepancy grid ends at
$T-\Delta t$, following the FBCIR reference code. `t` is therefore the anchor
$T$, while the reported interval ends at $T-\Delta t$.

Exact and ratio results are labelled `layer_cake_on_O(dt)_T_minus_dt_grid`
and `upper_ratio_on_O(dt)_T_minus_dt_grid`. Unlike the forward ratio, the
backward ratio is a plain L1 grid sum (Andreou & Chen, 2026, eq. G.14),
following the FBCIR reference code's convention; `quadrature` switches either
direction's ratio to the other rule.

Low-signal anchors, those whose peak finite-lag divergence $M$ is small, are
masked by the default `min_M = "auto"` rule rather than reported as inflated
lengths; the default `masked_value = "na"` keeps
the masking visible, and `masked_value = "zero"` restores the published zero
convention. Interpret CIRs jointly with the metric (Andreou & Chen, 2026,
Remarks B.4 and C.4). The ensemble counterpart of the forward table is in
Section 6.


## 4. Conditional ACI

With several observed effects, the conditional question is
$y \to x_A \mid x_B$: keep $x_B$'s values in the dynamics but give its
innovations no assimilation weight. That is `strategy = "inflate"`, used
below; the default, `strategy = "prescribed_forcing"`, instead substitutes the
non-target channels as known forcing and drops them from the observed block.
Here we use the model's recorded reference partition; choosing the `hidden`
partition of a benchmark model is a separate modelling decision:

```{r conditional}
me <- model_enso6(hidden = "tau")   # tau (wind bursts) hidden; u, hW, TC, TE, I observed
se <- simulate(me, seed = 12, T = 1.5, dt = 5e-3, burn_in = 0.5)
init_cond <- list(mean = 0, cov = diag(1, me$l))
a_all  <- aci(me, se$obs, init = init_cond)
a_cond <- aci(me, se$obs, init = init_cond,
              nontarget = nontarget(blocks = me$meta$conditioning_obs_idx,
                                    strategy = "inflate"))
plot(a_all$t, a_all$aci, type = "l", xlab = "t", ylab = "ACI",
     ylim = range(a_all$aci, a_cond$aci),
     main = "Total vs conditional influence of the wind bursts")
lines(a_cond$t, a_cond$aci, col = 2)
legend("topright", c("tau -> all observed", "tau -> TC,TE,I | u,hW"),
       col = 1:2, lty = 1, bty = "n")
```


## 5. Significance and validation

Two independent checks guard against wishful reading; both are package
extensions, recorded as such on each result (`source_status`). The
structural check asks whether a queried direction is *nil*: whether the model's
direct coupling block reaches it, and whether the empirical metric agrees. On the coupled
dyad both flags come back `FALSE`, meaning coupling is present and the metric
sees it; `TRUE` would flag the direction as nil. `direct sensitivity` is the
largest direct model coupling from the queried cause into the queried effect
along the record, and `empirical peak` is the largest metric value on it.
Supplying `obs` runs the empirical half on the series in hand; omitting it
simulates a fresh one:

```{r nilcheck}
nil_causality_check(m, sim$obs, direction = list(cause = 1, effect = 1))
```

The statistical check asks whether the *data* could have produced this
evidence without coupling. It is a likelihood-ratio parametric bootstrap
against the decoupled null (the candidate with its coupling removed), with
a flagged linearized fallback when most null simulations diverge:

```{r surrogate}
nil_surrogate_test(m, sim$obs, B = 19, seed = 2,
                   init = list(mean = 0, cov = diag(1, 1)))
```

`B = 19` keeps the vignette cheap; 0.05 is the floor of the p-grid at that
`B`. When the model was learned from the same series (via `aci_fit()`,
Section 7), the test is not out of sample: the coupled model was tuned to
these observations, and neither model is re-estimated on the surrogate
replicates, so the null ratios carry no matching fitting advantage. Small
p-values are optimistic there.


## 6. The ensemble engine

For systems outside the CGNS class (or as an independent cross-check), the
ensemble Kalman-Bucy filter/smoother needs only pointwise drift and diffusion
evaluations, not the conditional Gaussian split; it does require independent
observation and signal noise, and a member-independent signal diffusion. The
smoother **must reuse the forward Wiener increments** (Jiang et al., 2026).

### Filter and smoother

The package records and checks those increments before the backward pass.
The filter is started from the same prior as the closed-form run so the two
smoothers are comparable:

```{r ensemble}
ef <- enkbf(model = m, sim$obs, m = 40, seed = 4, ic_sampler = init)
es <- enkbs(m, ef$path, ef$noise)
em <- rowMeans(es$members[1, , ])
plot(sim$obs$t, smoo$mean[, 1], type = "l", xlab = "t", ylab = "y",
     ylim = range(smoo$mean[, 1], em),
     main = "Closed-form vs ensemble smoother (m = 40)")
lines(sim$obs$t, em, col = 4, lty = 2)
legend("topright", c("closed form", "EnKBS mean"), col = c(1, 4),
       lty = 1:2, bty = "n")
```

For high-dimensional systems (`model_l96()`), localization
(`localization_spec()`) tapers spurious long-range sample correlation and
inflation counters spread collapse in small ensembles; both are state
estimation devices, and the ACI metric and the ensemble CIR still require the
ensemble size to exceed the hidden dimension.

### The ensemble CIR

The **forward ensemble CIR** is reached through `aci()`. It reuses one causal
EnKBF run and repeats the EnKBS backward pass at each observation horizon,
building the complete versus lagged smoother family of Andreou & Chen (2026).
Jiang et al. (2026) use it for their forward CIR experiment. The construction
is $O(N^2)$ in time points, for both work and storage, with full-dimensional
Gaussian ensemble-moment KL. It is not run here: on this 1201-point series it
takes roughly ten minutes. Try it on a short series, and check results across
seeds and ensemble sizes:

```{r ensemble-cir, eval=FALSE}
ae  <- aci(model = m, obs = sim$obs, engine = "ensemble", m = 50, seed = 4,
           keep = "table")
cef <- forward_cir(ae)
# equivalently: forward_cir(ensemble_lag_table(m, sim$obs, m = 50, seed = 4))
```


## 7. Learning models from data

The learning layer supplies the components `aci_fit()` is built from:
polynomial function libraries (`cgns_library()`), structure selection by
**causation entropy** (`causation_entropy()`, thresholded through
`aci_fit(ce_threshold = )`), and constrained maximum likelihood including
cross-equation **energy conservation** constraints of the dyad/triad family.
`learn_model()` itself is the partial-observation path:

```{r discovery}
lib <- cgns_library(k = 1, l = 1, degree = 2)
lm_ <- learn_model(m, sim$obs, lib, n_samples = 10, seed = 3,
                   energy_pairs = list(c(obs = 1, hid = 1)), init = init)
round(c("dx1: x1*y1" = lm_$equations$dx1$coef[["x1*y1"]],
        "dy1: x1^2"  = lm_$equations$dy1$coef[["x1^2"]]), 3)
```

The two coefficients sum to zero because `energy_pairs` imposes that equality
as a hard constraint on the joint refit: the exact zero is the constraint being
met, not a recovery check. The recovery check is the magnitude itself, against
the dyad's true `gamma = 2`; the overshoot is the documented one-pass bias on
quadratic terms, which is why partial-observation fitting carries the
experimental label (see `?learn_model` and `vignette("vignette-1-intro")`).


## 8. The extreme-event workflow

The Moser et al. (2026) event workflow links the ACI metric to intermittent
extreme events; the dyad's antidamping bursts are the example used here. The
package covers this workflow only in part; see the limitations and roadmap
section of the README. `threshold = "event_peak_quantile"` is the Moser
dyad convention (the raw trajectory quantile route serves the pathways-style
workflow instead):

```{r extremes, fig.height=3.2}
sl <- simulate(m, seed = 5, T = 14, dt = 2e-3, burn_in = 2)
ev <- detect_events(sl$obs, component = 1, threshold = "event_peak_quantile",
                    q = 0.95)
plot(sl$obs$t, sl$obs$x[, 1], type = "l", xlab = "t", ylab = "x",
     main = sprintf("%d detected events", nrow(ev)))
abline(v = ev$t_star, col = 2, lty = 3)
```

The rest of the event layer:

* `event_onset()`: precursor times.
* `event_influence()`: per-event forward and backward influence windows.
* `event_stats()`: event-locked hidden-state composites against climatology.
* `classify_events()`: mechanism clustering, fed by the model-specific
  `features_pathways()` or `features_topographic()`; neither applies to the
  dyad.
* `sensitive_directions()`: the hidden-state directions along which
  event-conditioned posteriors depart most from climatology (a multistart
  numerical search, not a certified optimum).

In `event_influence()`, the windows combine a last-exceedance objective
(Andreou & Chen, 2026) with event anchors (Moser et al., 2026). Both windows
are package syntheses: the forward value agrees with the Moser et al. (2026)
definition when the finite-lag divergence decreases monotonically and differs
on non-monotone rows; the returned `forward_bound` and `backward_bound`
columns carry the CIR labels.
