---
title: "Assimilative causal inference on the nonlinear dyad model"
output:
  rmarkdown::html_vignette:
    math_method: mathml
vignette: >
  %\VignetteIndexEntry{Assimilative causal inference on the nonlinear dyad model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

Assimilative causal inference (ACI) quantifies, step by step, how strongly an
observed signal carries information about an unobserved component of the same
system. The idea rests on data assimilation. A forward filter estimates the
hidden component from the observed path seen so far, a backward smoother
estimates it from the whole observed path, and the causal-information metric
reads off how much the future of the observed signal sharpens the estimate at
each moment. The method reimplemented here is due to Andreou, Chen and Bollt
(2026).

This vignette walks through the whole workflow on the nonlinear dyad model
with intermittent extreme events, the conditional Gaussian nonlinear system
that the reference implementation uses as its worked example.

```{r library}
library(aciR)
```

## The nonlinear dyad model as a conditional Gaussian nonlinear system

A conditional Gaussian nonlinear system (CGNS) pairs an observed process $x$
with an unobserved process $y$ whose statistics, conditional on the observed
path, are Gaussian and available in closed form. The dyad model is the CGNS
governed by the pair of stochastic differential equations

$$
\mathrm{d}x = (-d_x\,x + \gamma\,x\,y + F_x)\,\mathrm{d}t
  + \sigma_x\,\mathrm{d}W_x,
$$

$$
\mathrm{d}y = (-d_y\,y - \gamma\,x^2 + F_y)\,\mathrm{d}t
  + \sigma_y\,\mathrm{d}W_y .
$$

The observed process $x$ couples to the hidden process $y$ through the
state-dependent term $\gamma\,x\,y$. That feedback is what makes $x$
intermittently extreme, since the bursts in the observed signal are driven by
the hidden component. The drift of $y$ is linear in $y$, which is the
structural condition that keeps the conditional statistics of $y$ Gaussian and
gives the filter and smoother their closed forms.

The constructor `aci_dyad_model()` builds the model object with the reference
parameters.

```{r model}
model <- aci_dyad_model()
model
```

## The causal question

The dyad model is built so that the hidden $y$ drives the observed $x$. The
question ACI answers is not merely whether that coupling exists but *when* it
is informationally active. At which moments does watching the future of $x$
tell us something new about $y$ that its past alone did not?

ACI answers this by contrasting two estimates of the hidden component. The
filter posterior of $y(t)$ conditions on the observed path up to time $t$. The
smoother posterior conditions on the whole observed path, past and future. The
causal-information metric is the relative entropy (Kullback-Leibler
divergence) of the smoother posterior from the filter posterior at each step,
and it is non-negative. Where it is large, assimilating the future of the
observed signal has revised the estimate of $y$ sharply, which marks a step of
strong causal coupling.

## Simulate a realisation

`aci_simulate()` integrates the model's equations by an Euler-Maruyama step and
returns the observed process $x$ together with the hidden process $y$. Because
the hidden $y$ is known in a simulation, it can be laid alongside the inferred
estimates as a ground truth. That check is never available with real data.

```{r simulate}
sim <- aci_simulate(model, n = 5000, seed = 333)
head(sim)
```

## Run assimilative causal inference

`aci()` is the single entry point. It assembles the conditional Gaussian
components from the model and the observed signal, runs the forward filter and
the backward smoother, and scores each step with the causal-information metric.

```{r infer}
fit <- aci(sim$x, model)
fit
```

The returned `aci` object holds the model, the observed signal, the `filter`
and `smoother` statistics (each a list of the posterior `mean` and `cov` of the
hidden component) and the causal-information metric in `aci`.

## The filter and the smoother

The filter and the smoother both estimate the hidden $y$, but from different
information. The filter uses only the past of the observed signal, so its
estimate is what a real-time observer would hold. The smoother uses the whole
record, so it is the retrospective estimate. Plotting both against the true
hidden path shows the smoother tracking $y$ more closely, most visibly where
the two estimates part company.

```{r filter-smoother, eval = requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)

estimates <- data.frame(
  t = fit$t,
  truth = sim$y,
  filter = fit$filter$mean,
  smoother = fit$smoother$mean
)

ggplot(estimates, aes(x = t)) +
  geom_line(aes(y = truth, colour = "true y"), linewidth = 0.4) +
  geom_line(aes(y = filter, colour = "filter mean"), linewidth = 0.4) +
  geom_line(aes(y = smoother, colour = "smoother mean"), linewidth = 0.4) +
  scale_colour_manual(
    values = c(
      "true y" = "grey55",
      "filter mean" = "#d95f02",
      "smoother mean" = "#2c7fb8"
    )
  ) +
  labs(
    x = "time",
    y = "hidden component y",
    colour = NULL,
    title = "Filter and smoother estimates of the hidden component"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

## The causal-information metric

The metric itself is a time series. Laying it under the observed signal shows
where the two align. The spikes in causal information sit at the intermittent
bursts of the observed $x$, which is exactly where the hidden $y$ is driving
the observed process hardest and where the future of $x$ is most informative
about $y$. The two series get a panel each because they are incommensurable.
The signal is in its own units and the metric is in nats, so it is the timing
that aligns between the panels, never the magnitudes.

```{r metric, fig.height = 5, eval = requireNamespace("ggplot2", quietly = TRUE)}
series <- rbind(
  data.frame(t = fit$t, series = "observed x", value = fit$x),
  data.frame(t = fit$t, series = "causal information (nats)", value = fit$aci)
)
series$series <- factor(
  series$series,
  levels = c("observed x", "causal information (nats)")
)

ggplot(series, aes(x = t, y = value, colour = series)) +
  geom_line(linewidth = 0.4, show.legend = FALSE) +
  facet_grid(rows = vars(series), scales = "free_y") +
  scale_colour_manual(
    values = c(
      "observed x" = "grey40",
      "causal information (nats)" = "#2c7fb8"
    )
  ) +
  labs(
    x = "time",
    y = NULL,
    title = "Causal information under the observed dyad signal"
  ) +
  theme_minimal()
```

The `summary()` method reports the metric alongside two diagnostics that say
whether the numerical assumptions held. The smallest posterior covariances show
how close the explicit Euler scheme came to losing positivity. The
terminal-identity residual is zero analytically, so it measures accumulated
numerical error.

```{r metric-summary}
summary(fit)
```

Every series is available in one frame with `as.data.frame()`, which is the
tidy starting point for a figure or an export.

```{r as-data-frame}
head(as.data.frame(fit), 3)
```

The peaks carry a bounded reading. The metric says that *under this model* the
future of the observed signal is informative about the hidden component at
those moments. It does not test the model, and it does not identify the effect
of an intervention. Here the dyad model is also the process that generated the
data, so the reading holds without further argument. On real data the model is
an assumption rather than a fact, and the article *Assumptions and
interpretation* sets out which claims an ACI peak supports and which it does
not.

The quantity plotted above is a relative entropy between two posteriors over
the same instant, the one informed by the whole observed record and the one
informed only up to that instant.

$$
\mathrm{ACI}(t)
  = \mathrm{KL}\left( p(y_t \mid x_{0:T}) \,\middle\|\, p(y_t \mid x_{0:t}) \right).
$$

The Gaussian closed form this reduces to, and the identities it is graded
against, are in the companion article *Validation and the independent oracle*.

One scope note closes the walk-through. This vignette covers the
causal-information metric, which answers *how much* the future of the observed
signal says about the unobserved state. The paper defines a second quantity,
the causal influence range, which answers *how far ahead* one must look before
that answer stops changing. That quantity is implemented as `aci_cir()` and
graded against the same reference. It is not covered here because it asks a
different question rather than extending this one, and `?aci_cir` is where to
begin on it.

## The general components path

`aci()` is a convenience layer over three core functions that work on any
conditional Gaussian nonlinear system rather than on the dyad model alone.
Each takes a components list, the per-step coefficients of the governing
equations, so the same core serves a custom system once its components are
supplied. For the dyad model, `aci_dyad_components()` builds that list from the
observed signal, and the filter, the smoother and the metric can be called in
turn.

```{r components}
comp <- aci_dyad_components(sim$x, model$parameters)
filt <- aci_filter(sim$x, comp, dt = 0.001, mu0 = 2, R0 = 0.1)
smooth <- aci_smoother(sim$x, comp, dt = 0.001, filt)
metric <- aci_metric(filt, smooth)

summary(metric)
```

This is the general form of the workflow. For a conditional Gaussian nonlinear
system of one's own, `aci_cgns_model()` builds the model object and `aci()`
runs the same three steps.

## References

Andreou, M., Chen, N. and Bollt, E. (2026). Assimilative causal inference.
*Nature Communications*, 17, 1854.
<https://doi.org/10.1038/s41467-026-68568-0>
