---
title: "Assumptions and interpretation"
output:
  rmarkdown::html_vignette:
    math_method: mathml
vignette: >
  %\VignetteIndexEntry{Assumptions and interpretation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## What this article covers

The other vignette shows a large peak in the causal-information metric and
describes it as a moment of strong causal coupling. That reading is correct
only under conditions the software cannot check. This article states those
conditions, so that a reader can say when the method applies, when the software
should refuse to run, and what an ACI peak does and does not support.

```{r library}
library(aciR)
```

## What the metric is

At each time step aciR reports the relative entropy of the smoother posterior
of the unobserved component from its filter posterior. The filter posterior is
the law of the unobserved state given the observed path so far. The smoother
posterior is its law given the whole observed path. The metric therefore
answers one question, step by step. How much does knowing the *future* of the
observed signal change what we believe about the unobserved component *now*?

This is an information-theoretic quantity, and it is non-negative by
construction. It has units of nats. A value of zero means the future of the
observed signal is uninformative about the hidden state at that step, and a
large value means it is highly informative.

## The estimand, stated precisely

The quantity is a property of a **model-conditional posterior pair**, not of
the world. Written out, aciR estimates

$$
\mathrm{ACI}(t) \;=\; \mathrm{KL}\!\left(
  p(y_t \mid x_{0:T}, \mathcal{M}) \;\|\; p(y_t \mid x_{0:t}, \mathcal{M})
\right),
$$

where $\mathcal{M}$ is the conditional Gaussian nonlinear system you supplied,
$x$ is the observed process, $y$ is the unobserved one, and $T$ is the end of
the record. Every claim the metric supports is conditional on $\mathcal{M}$.
Change the model and the metric changes, with no warning from the software,
because a different model is a different question.

The direction is fixed and asymmetric. The metric measures information flowing
from the observed process to the estimate of the unobserved one. It is not a
test of whether $y$ causes $x$, and reversing the roles of the two processes
requires a different model, not a different reading of the same output.

## The assumptions

### Model correctness

The method assumes the supplied system is an adequate description of the data.
The filter and smoother are exact for the model, so all of the approximation
lives in the gap between the model and reality. There is no goodness-of-fit
test in this package, and a badly specified model will still produce a smooth,
plausible metric trace. This is the assumption most likely to be wrong and the
least likely to announce itself. The check has to come from outside the metric.
One route available here is to draw realisations from the supplied model with
`aci_simulate()` and compare their statistics against the observed record
before reading anything into a peak.

### Conditional Gaussian structure

The unobserved process must enter its own drift linearly and the noise must not
depend on the unobserved state. That is what makes the conditional statistics
Gaussian in closed form. Systems outside this class are not approximated by
aciR, they are simply not expressible in it. Such a system needs a filtering
method that does not rely on closed-form conditional statistics, which is
outside the scope of this package.

### Known parameters

The coefficients are taken as given. aciR does not estimate them and does not
propagate uncertainty in them into the metric. Where the parameters came from a
fit, the reported metric is conditional on a point estimate, and its true
uncertainty is wider than anything the package reports by an amount the package
cannot see. The route to that uncertainty is to rerun the analysis over a set
of parameter values drawn from the fit and to report the spread of the metric
across them.

### State dimension

The numerical core takes vector-valued states, an observed process of any
dimension and an unobserved one of any dimension, with matrix coefficients that
may vary in time. The filter, smoother and metric all dispatch on the shape of
the components you supply.

The fixed-lag online smoother and the causal influence range built on it
dispatch the same way, so a vector system is carried end to end.

One path is graded more weakly than the rest. The matrix noise-cross-covariance
terms are implemented, but the reference implementation contains no model that
exercises them (its ENSO script says so in as many words), so they are graded
by collapse onto the scalar path and by analytic fixed points rather than
against the authors' numbers. Grading them against the authors' numbers would
require an upstream model carrying a non-zero cross-covariance, and no such
model exists. The companion article *Validation and the independent oracle*
records the exact scope of that grade.

### Conditional questions

With more than one observed component, `aci_conditional()` asks what one of
them says about the unobserved state given that the others are also watched.
This changes the **estimand**, not the arithmetic. The result answers a
different question rather than answering the same one more precisely. The
non-target components are not removed, and they still drive the system. What
changes is that the filter stops assimilating them, which is not the same as
never having observed them.

### Self-drift measurable with respect to the observed path

The self-drift `L_y` of the unobserved component may be constant in time, as in
the reference dyad model, or it may vary, as in the noisy predator-prey model
where the latent population's growth rate is set by the population being
watched. What it may *not* do is depend on the unobserved component itself.
Such a dependence takes the system outside the conditional Gaussian class and
leaves the closed-form recursions without a posterior to propagate.

### Regular, complete sampling

The recursions integrate a fixed step. The observed signal must be complete and
sampled on a regular grid. There is no missing-observation contract. aciR
rejects a signal containing `NA` or an infinite value rather than interpolating
one, and rejects an irregular `time` grid rather than approximating it.
Completing or regularising a record is therefore the caller's step, and the
choice made there should be reported alongside the result.

```{r reject-missing, error = TRUE}
model <- aci_dyad_model()
aci(c(1, NA, 3), model)
```

### Discretisation

The filter and smoother are integrated with a first-order Euler scheme. Two
consequences matter in practice. Identities that hold exactly in continuous
time hold only to order `dt` here, and an explicit Euler step that is too large
for the system can drive the posterior covariance out of its domain even when
the model is perfectly admissible. In that state the relative entropy has no
meaning, so aciR stops and names the step rather than returning a number. The
remedy in both cases is a smaller `dt`, and the next section sets out how to
check that the value chosen is small enough.

```{r reject-unstable, error = TRUE}
x <- aci_simulate(model, n = 200, seed = 1)$x
aci(x, model, dt = 1)
```

## Sensitivity you should report

Three inputs are choices rather than data, and a result that changes materially
across reasonable values of them is a result about the choice.

* **The time step `dt`.** Halve it and confirm the metric is stable. The
  package's own tests show the discretisation residual falls at second order in
  the step, so a halving that changes your conclusion means you were not yet in
  the convergent regime.
* **The initial filtered covariance `R0`.** Its influence decays as the filter
  forgets its initial condition, but early steps are sensitive to it.
* **The initial filtered mean `mu0`.** Same reasoning.

`summary()` surfaces the two diagnostics that bear on this. The smallest
posterior covariances say how close the run came to losing positivity. The
terminal-identity residual measures accumulated numerical error against a
quantity that is zero analytically.

```{r summary}
fit <- aci(aci_simulate(model, n = 5000, seed = 333)$x, model)
summary(fit)
```

## What an ACI peak does and does not support

A peak supports one claim. **Under the supplied model**, the future of the
observed signal is unusually informative about the unobserved component at that
moment. For a well-specified model this is a statement about the mechanism, and
it is the statement the method was designed to make.

A peak does not support any of the four claims below.

1. **That the model is right.** The metric is computed conditional on the
   model, so it cannot test it. A misspecified system yields a confident metric
   about a mechanism that does not exist.
2. **That an intervention would have an effect.** The estimand is defined by
   conditioning, not by intervening. Nothing here identifies a causal effect in
   the interventional sense, and no assumption made by this package licenses
   that step.
3. **That causality has been discovered from data alone.** The causal content
   comes from the model, which encodes which process drives which. The data
   locate *where in time* that assumed structure carries information. Structure
   in, structure out.
4. **That a larger value in one system exceeds a larger value in another.** The
   metric is in nats and is comparable across time within one model and one
   record. Comparing across models, parameterisations or sampling rates
   compares the conditioning as much as the coupling.

In short, assimilative causal inference reads *when* a mechanism you have
specified transmits information. It does not establish *whether* that mechanism
exists. The first of those is a useful thing to know, and it is narrower than
the phrase "causal inference" suggests to a reader arriving from the
potential-outcomes literature. State the distinction explicitly in anything
published from the method.

## Reading the causal influence range

The range is a **last-exit time, not a duration**. It is the last moment at
which the divergence still exceeds a threshold, which is at least the amount of
time spent above that threshold and is strictly more whenever the sequence
rises again after falling.

That distinction decides how the two objective ranges relate. `objective`, the
efficient integral, and `objective_exact`, the average of the subjective ranges
over the threshold grid, are the same functional only when the comparison
sequence decreases with lag. `summary()` reports how often that holds; where it
does not, seeing one below the other is a definitional gap rather than a
numerical defect in either.

A time marked `censored` is a **lower bound, not a missing value**. Its range
consumed more of the record than the retained margin allows, so the record
cannot settle it. The record does support the statement that the range is at
least this long, and that statement is what is returned. Only a record long
enough for the range to fit inside it will resolve the value.

`aci_conditional()` carries its own estimand note; see `?aci_conditional` for
what conditioning does and does not mean.

## Diagnostics to report

For a result intended for publication, report at least the following, all of
which the package makes available.

* The full model specification and the provenance of every parameter, stating
  which were estimated and from what.
* `dt`, `mu0`, `R0`, and the sensitivity of the conclusion to each.
* The smallest posterior covariance over the run.
* The sampling rate of the observed record and any preprocessing used to make
  it regular and complete.
* The package version, since the metric depends on the numerical core.

## Further reading

The method is due to Andreou, Chen and Bollt (2026),
[doi:10.1038/s41467-026-68568-0](https://doi.org/10.1038/s41467-026-68568-0).
The companion article *Validation and the independent oracle* records how this
implementation is graded and what that grade covers.
