---
title: "Validation and the independent oracle"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation and the independent oracle}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The problem this addresses

A test suite written by the author of the code it tests can only prove the code
is self-consistent. If the author misread an equation, the test encodes the
same misreading and passes. For a package whose entire purpose is to reproduce
a published numerical method, self-consistency is close to worthless: the
question is not whether aciR agrees with itself, but whether it agrees with the
method.

Every numerical claim aciR makes is therefore graded against a reference the
package did not produce. This article records what those references are, how
they were generated, what each one covers, and -- the part most easily
overlooked -- what each one does *not* cover.

```{r library}
library(aciR)
```

## The three oracles

aciR uses three, because no one of them is sufficient.

The quantity all of them grade is, at each step, the relative entropy of the
smoother posterior $\mathcal{N}(m_s, R_s)$ from the filter posterior
$\mathcal{N}(m_f, R_f)$, which for scalar Gaussians has the closed form

$$
\mathrm{KL}\!\left(\mathcal{N}(m_s, R_s) \,\|\, \mathcal{N}(m_f, R_f)\right)
= \frac{1}{2}\left[
    \frac{(m_s - m_f)^2}{R_f}
    + \frac{R_s}{R_f} - 1
    - \log\frac{R_s}{R_f}
  \right],
$$

together with the four posterior series that feed it. The first term is the
signal part, driven by the shift between the two means, and the remainder is
the dispersion part, driven by the covariance ratio; `aci_metric()` evaluates
the dispersion part in a cancellation-resistant form, which is one of the
things the oracles hold it to.

### 1. The dyad fixture, from the authors' reference implementation

The flagship oracle. A MATLAB harness reproduces the deterministic core of
`dyad_interaction_model.m` from the reference implementation at
[github.com/marandmath/ACI_code](https://github.com/marandmath/ACI_code)
(MIT licence, commit `733c49f`), using that reference's own seed and
parameters. It writes two files: the observed signal, and the expected filter
mean, filter covariance, smoother mean, smoother covariance and ACI metric at
every hundredth step.

aciR then runs its *own* filter, smoother and metric on that signal and must
reproduce all five series. The comparison is at 301 sampled indices of a
30,001-step trajectory, and the gate is a maximum absolute error below `1e-6`.

The observed agreement is far tighter than the gate:

| Oracle | Maximum absolute error | Gate |
|---|---|---|
| Dyad | `4.574119e-14` | `1e-6` |
| Cross-noise | `1.376677e-14` | `1e-6` |

Both sit at the floating-point noise floor of a 30,001-step recursion. This is
not agreement within tolerance; it is agreement to the last few bits the
arithmetic can carry.

### 2. The cross-noise fixture, and why it had to exist

The dyad oracle is strong, but its scope is narrower than it first appears, and
the gap is instructive.

Every scalar model in the reference implementation -- the dyad and the noisy
predator-prey model alike -- sets the noise decomposition so that the noise
cross-covariance is zero. The filter's update carries a term
`aux = S_yoS_x + R * L_x`, and the smoother carries `A_j` and a transport term,
all of which involve that cross-covariance. On every run of the dyad fixture
those terms *execute*, and are multiplied by zero.

The consequence is worth stating plainly. Line coverage sees the arithmetic run
and reports it covered. The oracle sees the outputs agree and reports them
graded. Neither is wrong, and yet the cross-covariance terms had never been
checked against anything -- while `aci_cgns_model(S_yoS_x = ...)` exposes them
to users as a supported path. A high coverage figure on a graded path had been
manufacturing confidence in an ungraded path beside it.

So aciR ships a second harness, `aci_oracle_cross.m`, which is the reference
dyad with correlated noise switched on. It has no upstream counterpart, because
upstream has no scalar model that exercises these terms. It is therefore a
second independent implementation of the published equations in a different
language and runtime. Agreement with it refutes a transcription error on the R
side over the full transient. It is *not* an authors'-reference grounding, and
the manifest says so.

### 3. The analytic identities, which depend on no implementation

The primary grounding for the cross-noise path is not either fixture. It is
algebra.

For a constant-coefficient system the stationary filtered covariance is the
positive root of an algebraic Riccati equation, obtained by setting the
covariance derivative to zero:

$$
L_x^2 R^2 \;-\; 2\left(L_y S_{xx} - S_{yx} L_x\right) R
\;-\; \left(S_{yy} S_{xx} - S_{yx}^2\right) \;=\; 0 .
$$

Started at that root with a non-zero cross-covariance, the recursion must stay
there to machine precision, because the fixed point of the differential
equation is exactly the fixed point of its Euler map. This oracle is derived
from the governing equations and is independent of aciR *and* of MATLAB. The
package's tests also check the smoother's analogous fixed point, the terminal
identity, and the zero-information limit.

The zero-information test deserves a note, because it is where a naive test
would have been wrong. With no coupling and no cross-covariance the observed
process is independent of the unobserved one, so the smoother must equal the
filter and the metric must be zero. That identity is exact in continuous time
only. Under an explicit Euler scheme the forward and backward sweeps do not
invert one another exactly, so asserting equality would assert a property of
the mathematics against an approximation of it. The test instead asserts that
the residual vanishes at second order in the step, which is the true claim and
a stronger one: it checks the *rate*, not just the size.

## Provenance and the manifest

Trusting a committed fixture requires knowing where it came from and being able
to tell an intentional change from an accidental one.
`inst/extdata/oracle-manifest.yml` records, for each fixture, the upstream
repository and immutable commit, the licence and copyright holder, the MATLAB
release, the exact generating command, the parameters and seed, the sampled
indices, both an MD5 and a SHA-256 hash, the measured error -- and a `grades`
and `does_not_grade` field.

```{r manifest}
manifest <- system.file("extdata", "oracle-manifest.yml", package = "aciR")
cat(paste(readLines(manifest)[1:12], collapse = "\n"))
```

The recorded MATLAB release is grounded by reproduction rather than by
recollection: the harness was rerun on that release into a scratch directory
and both files reproduced byte-for-byte, with identical hashes to the committed
copies. A test checks the shipped bytes against the recorded hashes on every
run, so a fixture cannot drift silently.

## What is graded, and what is not

An oracle grades the paths its parameters exercise and no others. Two lists,
because the distinction that matters is not whether a thing works but whether
the evidence for it comes from the authors.

### Graded, with the scope named

* **A time-varying self-drift**, in both causal directions, against the noisy
  predator-prey model -- the system whose latent damping is set by the observed
  population. The dyad and cross fixtures cannot reach this path, since both
  pin the self-drift at a constant.
* **The noise cross-covariance of the online smoother and the causal influence
  range**, against the `cir_cross` fixture. This is a second independent
  implementation rather than the authors' own numbers: every scalar model
  upstream sets the cross-covariance to zero, so no reference fixture can reach
  these terms.
* **Vector filter, smoother and metric**, against the ENSO fixture and against
  block-diagonal algebraic Riccati fixed points solved in closed form, which
  depend on neither implementation.
* **The vector online smoother and causal influence range**, by collapse onto
  the graded scalar path at every lag and by an independent transcription of
  equations (3.5) to (3.7) in full generality. Not against the authors' ENSO
  scripts.

The matrix noise-cross-covariance terms deserve their own sentence. They have
**no upstream counterpart at all**: every scalar model in the reference sets
the cross-covariance to zero, and the ENSO scripts state in as many words that
the cross-interaction terms are absent. The analytic identities are the primary
grounding there, and a second implementation refutes a transcription error
without being an authors'-reference grounding.

### Still open

* **Simulation.** `aci_simulate()` draws from R's generator, which differs from
  MATLAB's, so a simulated path does not correspond to the reference path even
  at a matching seed. The oracle grade runs on the MATLAB signal, never on a
  fresh simulation. Simulation is checked for reproducibility and for its
  contract, not against a reference path.
* **`aci_conditional()`** against the one reference script that has conditional
  ACI enabled. The construction is tested; the numbers are not compared.
* **The predator-prey causal influence range.** Filter, smoother and metric are
  graded in both directions; the range is not.
* **Four of the five ENSO scripts**, whose workspaces are captured and
  validated but whose extracted kernels are not yet gated.
* **Model adequacy.** No oracle can grade this, and none pretends to. Whether
  the system you supplied describes your data is a scientific question, treated
  in the companion article *Assumptions and interpretation*.

## Reproducing the grade

The oracle tests run from the installed package fixtures. They never skip, on
any platform, in any check environment -- a validation gate that can skip is not
a gate, and an earlier version of these tests could skip when run outside the
development tree.

```{r reproduce, eval = FALSE}
testthat::test_local(system.file(package = "aciR"))
```

Regenerating the fixtures needs MATLAB and the harnesses in the project's
`oracle/` directory. The commands are in the manifest.
