---
title: "Assimilative causal inference (ACI) in R"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml.js"
    pandoc_args: ["--include-in-header=mathjax-config.html"]
vignette: >
  %\VignetteIndexEntry{Assimilative causal inference (ACI) in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 3.6, collapse = TRUE,
                      comment = "#>")
library(acir)
set.seed(1)
```

## 1. The question

Assimilative causal inference (Andreou, Chen and Bollt, 2026) treats
causality as an inverse problem. Given a model of how an observed process
$x$ and a hidden process $y$ evolve together, it asks whether the future of
$x$ carries information about $y_t$ that the past of $x$ did not. If it
does, $y$ drives $x$, and the amount of new information is the strength of
that influence at time $t$. The answer is computed along one trajectory,
so it describes causality as it happened rather than on average.

The causal influence range (Andreou and Chen, 2026) extends the question
from strength to duration: for how long does an influence planted at time
$t$ persist, and how far back did the precursors of an event reach. This
release computes the forward range.

## 2. The model comes first

In a regression workflow the model is the output: `lm()` is handed data
and returns coefficients. Here the model is the input, a statement of how
the observed and hidden variables evolve and how noisy each is, supplied
before the question is asked. It can be a benchmark constructor such as
`aci_dyad_model()`, the coefficient blocks written directly with
`aci_model()`, or a full drift that `aci_model_from_affine()` splits for
you. *The closed-form ACI engine* describes the three routes. Learning the
model from data is not part of this release.

### Two reconstructions and one number

Given the model, two Gaussian reconstructions of the hidden state are
available in closed form:

* the **filter** $p(y_t \mid x_{\le t})$, what the past of $x$ says.
* the **smoother** $p(y_t \mid x_{\le T})$, what the past and the future
  of $x$ say together.

If the future sharpens the reconstruction, $y$ influenced $x$. The ACI
metric is the relative entropy between the two,

$$\mathrm{ACI}(t) = D_{\mathrm{KL}}\!\big(p^{\text{smoother}}_t \,\big\|\,
p^{\text{filter}}_t\big) \ge 0,$$

and it splits into a **signal** part, from disagreement between the two
means, and a **dispersion** part, from disagreement between the two
spreads.

## 3. A worked example

The model comes first, so the example simulates from it and holds the
hidden path aside to check the reconstruction against.

```{r quickstart-data}
truth <- aci_dyad_model(params = list(d_x = 1, gamma = 2, f_x = 1.5, s_x = 0.7,
                                      d_y = 1, f_y = 0.5, s_y = 2))
sim <- simulate(truth, seed = 8, t_end = 5, dt = 5e-3, burn_in = 1)
dat <- data.frame(t = sim$obs$t, x1 = sim$obs$x[, 1], y = sim$hidden[, 1])
head(dat, 3)
```

The engine takes an observation object rather than a data frame, so the
observed column goes through `observed_trajectory()`. That call is also
the way in for your own data: it derives the time step from the supplied
times and refuses a non-uniform grid.

```{r quickstart-obs}
ob <- observed_trajectory(dat$t, dat$x1, names = "x1")
ob
```

Supply the model, the observations and a prior on the hidden state.
`aci_filter()` and `aci_smoother()` return the two reconstructions, and
`aci()` runs the pair and scores every time point.

```{r quickstart-engine}
init <- list(mean = 0, cov = matrix(1, 1, 1))
filt <- aci_filter(truth, ob, init = init)
smoo <- aci_smoother(truth, ob, filter = filt)
res  <- aci(truth, ob, init = init)
```

Name the prior. Omitting `cov` falls back to a diffuse prior with a
warning, and the opening steps of such a run show the prior washing out
rather than anything causal.

## 4. Read the result

```{r quickstart-latent, fig.cap = "The smoother's reconstruction of the hidden path with a band of two standard deviations, against the path that was withheld.", fig.alt = "Line plot of the smoothed mean and its band over time, with the true hidden path overlaid."}
plot(smoo, truth = dat$y)
legend("topright", c("reconstruction, two standard deviations", "withheld truth"),
       col = c(1, 2), lty = 1, bty = "n")
```

```{r quickstart-metric, fig.cap = "The ACI metric along the record, with its signal and dispersion parts.", fig.alt = "Line plot of the metric over time with two dashed component curves."}
plot(res)
```

`print()` reports the peak of the metric and where it fell:

```{r quickstart-print}
res
```

## 5. The influence range

For how long does $y$ influence $x_1$? `aci_range()` answers per anchor
time:

```{r quickstart-cir, fig.cap = "The forward causal influence range at the reported anchor times.", fig.alt = "Line plot of the range against anchor time."}
fc <- aci_range(res, anchors = seq(1L, length(ob$t), by = 25L))
fc
head(data.frame(t = fc$t, tau = fc$tau, M = fc$M, status = fc$status), 4)
plot(fc)
```

Each row is one anchor time `t`. `tau` is the range in the model's time
units and `M` is the peak finite-lag divergence there. `anchors` chooses
which anchor times are reported, and only those rows are formed, which
matters because forming every row of a long record is quadratic work.

An anchor whose peak divergence is too small to divide by is masked rather
than reported as an inflated length, and the warning above says how many.
`status` carries that reading and three others:

```{r quickstart-status}
table(fc$status)
```

Here `r sum(fc$status == "censored")` of the `r length(fc$status)` anchors
come back `censored`, a property of a short record rather than of the
model: after an anchor late in a five-unit series there is not always
enough future left to resolve the influence it planted, and a censored
range is a lower bound. The decline of the range at the end of the plot is
the same finite-record effect. *The closed-form ACI engine* gives the full
status vocabulary and the conventions each number is computed under.

## 6. Scope of this release

What is here is the closed-form engine: models in the conditional Gaussian
nonlinear system (CGNS) class, for which the filter, the smoother and the
metric have exact solutions, and the forward influence range. Learning the
dynamics from data, the backward range, significance testing and the
ensemble engine for systems outside the class are not in this release.

*The closed-form ACI engine* (`vignette("vignette-2-advanced")`) documents
the machinery and its arguments. *Reproducing the reference MATLAB
codebase* (`vignette("vignette-3-matlab")`) states which numbers of the
authors' implementation this package reproduces, and under which
conventions. `?aci_references` lists the sources.

## References

Andreou, M., Chen, N. and Bollt, E. (2026). Assimilative causal inference.
*Nature Communications*, 17, 1854.
[doi:10.1038/s41467-026-68568-0](https://doi.org/10.1038/s41467-026-68568-0)

Andreou, M. and Chen, N. (2026). Bridging prediction and attribution:
identifying forward and backward causal influence ranges using
assimilative causal inference. arXiv:2510.21889.
[doi:10.48550/arXiv.2510.21889](https://doi.org/10.48550/arXiv.2510.21889)
