---
title: "Assimilative causal inference (ACI) in R"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Assimilative causal inference (ACI) in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 3.6, collapse = TRUE,
                      comment = "#>")
library(acir)
set.seed(1)
```

## 1. Answering causal questions

### Estimating causal effects
Causal inference techniques provide useful tools for answering questions about
causality using observational data, including tasks that require estimating the
size of treatment effects. Many such frameworks and methods exist for causal effect
estimation, including graphical modelling approaches for estimating the causal effect
of a treatment (the cause) on an outcome (the effect). Such approaches require 
observable causes and their effect(s), captured in the data, to identify a causal estimand.


### Tracing backward from effects to causes
Assimilative causal inference (ACI) (Andreou et al., 2026) is a complementary
approach that formulates causality as an inverse problem, using Bayesian data 
assimilation to trace causes backward from their observed effects. This approach
can detect if one variable (a latent cause) influences another (an observed effect),
and if so, for how long a causal influence persists.

### The causal influence range (CIR)

Andreou & Chen (2026) develop the causal influence range (CIR) for quantifying 
the duration of a causal effect in both forward and backward directions. 
This answers questions such as: (1) "for how long will a causal effect persist?" 
(forward CIR); and (2) "when did the causal precursors of this event emerge?" 
(backward CIR). 

## 2. Getting started with ACI in R

### A different way of thinking about modelling
ACI requires a dynamical model to identify instantaneous causal relationships (Andreou et al., 2026).
In a standard regression workflow the model is the output: `lm()` is handed
data and returns coefficients. Here the model is the input: a statement of how
the observed and hidden variables evolve together, and how noisy each is,
supplied before the causal question is asked. Fitting is not part of this
release. The model is written down: a benchmark constructor (`aci_dyad_model()`),
the coefficient blocks directly (`aci_model()`), or a full drift split for you
(`aci_model_from_affine()`); the advanced vignette describes all three routes.
Learning the model from data is reserved for a later release.

### The question ACI answers
Some variables can be observable while others remain hidden, and for a latent
process $y$ (the cause) influencing observed variables 
(the effect) in a time series: **is a causal influence detected, and if so, how 
strongly at each moment, and for how long?**

To answer this question, we can compare two reconstructions of $y_t$:

* the **filter** $p(y_t \mid x_{\le t})$: what past observations say;
* the **smoother** $p(y_t \mid x_{\le T})$: what past *and future* observations say.

If the future of $x$ sharpens our knowledge of $y_t$, then it confirms the existence
of a causal influence. 

The **ACI metric** is the relative entropy between the two posteriors,

$$\mathrm{ACI}(t) \;=\; D_{\mathrm{KL}}\!\big(p^{\text{smoother}}_t \,\big\|\,
p^{\text{filter}}_t\big) \;\ge\; 0,$$

computed *along one trajectory* (causality as it happened, not on average).


### Simulating example data

Since the model comes first, we can simulate from it and hold the hidden
truth aside to check the reconstruction against.

```{r quickstart-data}
truth <- aci_dyad_model(params = list(d_x = 1, gamma = 2, f_x = 1.5, s_x = 0.7,
                                  d_y = 1, f_y = 0.5, s_y = 2))
sim <- simulate(truth, seed = 8, T = 5, dt = 5e-3, burn_in = 1)
dat <- data.frame(t = sim$obs$t, x1 = sim$obs$x[, 1], y = sim$hidden[, 1])
head(dat, 3)
```

The engine takes an observation object rather than a data frame, so the
observed column goes through `observed_trajectory()` first. For your own data
that same call is the way in; it derives `dt` from the supplied times and
refuses a non-uniform grid.

```{r quickstart-obs}
ob <- observed_trajectory(dat$t, dat$x1, names = "x1")
ob
```

## 3. Run the engine

Supply the model, the observations and a prior on the hidden state.
`aci_filter()` and `aci_smoother()` return the two reconstructions of Section 2,
and `aci()` runs the pair and scores each time by the relative entropy between
them.

```{r quickstart-engine}
init <- list(mean = 0, cov = matrix(1, 1, 1))
filt <- aci_filter(truth, ob, init = init)
smoo <- aci_smoother(truth, ob, filter = filt)
res  <- aci(truth, ob, init = init)
```

Name the prior explicitly. Omitting `cov` falls back to a diffuse prior, with a
warning, and the opening steps of such a run report the prior washing out
rather than anything causal.

`aci()` also names the two parts of the metric: a **signal** term, from
disagreement between the posterior means, and a **dispersion** term, from
disagreement between the posterior spreads.

## 4. Plot and print the results

`plot()` visualises the reconstruction and the metric:

```{r quickstart-latent}
plot(smoo, truth = dat$y)
legend("topright", c("reconstruction ± 2 sd", "withheld truth"),
       col = c(1, 2), lty = 1, bty = "n")
```

```{r quickstart-metric}
plot(res)
```

`print()` reports the peak of the metric and where it fell:

```{r quickstart-print}
res
```

## 5. Obtain the CIR

The causal influence range (CIR) (*for how long does $y$ influence $x_1$?*) 
is obtained using `aci_range()`:

```{r quickstart-cir}
fc <- aci_range(res, anchors = seq(1L, length(ob$t), by = 25L))
fc
head(data.frame(t = fc$t, tau = fc$tau, M = fc$M, status = fc$status), 4)
plot(fc)
```

Each row is one anchor time `t`: `tau` is the range in the model's time units
and `M` is the peak finite-lag divergence there. `anchors` chooses which anchor
times are reported, and only those rows are formed, which matters because
forming every row of a long record is quadratic work. This release reports the
forward direction; the backward question of Section 1 is reserved for a later
release.

Anchors whose peak divergence is too small to divide by are masked rather than
reported as inflated lengths, and the warning above says how many. `status`
carries that reading and three others, one per anchor:

```{r quickstart-status}
table(fc$status)
```

Half of these anchors come back `censored`, which is a property of a short
record rather than of the model: after an anchor late in a five-unit series
there is not always enough future left to resolve the influence it planted,
and a censored range should be read as a lower bound. That is also what the
decline in $\tau_f$ at the end of the plot is.
`vignette("vignette-2-advanced")` gives the full status vocabulary and the
conventions each number is computed under.

## 6. What this release does not cover

The model here was written down and the hidden path was reconstructed from it.
Learning the dynamics from data instead, when the cause was never measured, is
the partial-observation problem, and it is not in this release. Neither is the
backward CIR, nor the significance machinery, nor the ensemble engine for
systems outside the CGNS class. What is here is the closed-form engine: models
in the conditional Gaussian nonlinear system (CGNS) class, for which the
filter, the smoother and the metric all have exact solutions.

We're well on our way to seeing what ACI can do. The underlying machinery is
covered in `vignette("vignette-2-advanced")`, and
`vignette("vignette-3-matlab")` covers the conventions under which the package
reproduces its reference MATLAB codebase.
