---
title: "The closed-form ACI engine"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{The closed-form ACI engine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 3.6, collapse = TRUE,
                      comment = "#>")
library(acir)
```

`vignette("vignette-1-intro")` runs the engine end to end. This one is
about the machinery underneath: how models are written, what the filter,
smoother and online-smoother arguments do, what happens when a covariance
leaves the positive-definite cone, which conventions the causal influence
range is computed under, and which of the conditional questions are
genuinely different estimands rather than different estimators.

Every convention below is a choice that changes the number reported. The
package states them on the result objects rather than leaving them to be
inferred, and this vignette is largely a guide to reading those statements.

Where a choice exists only to match the reference MATLAB codebase, it is
named here and worked through in `vignette("vignette-3-matlab")`, which also
covers the benchmark constructors' provenance and the kinds of evidence
behind the package's fidelity claims.


## 1. Models and observations

### The CGNS class

ACI's closed form rests on a structural property: **given the observed
path, the hidden dynamics are linear and Gaussian.** A conditional
Gaussian nonlinear system (CGNS) is written

$$\mathrm{d}x = \big[L_x(t,x)\,y + f_x(t,x)\big]\mathrm{d}t
  + \Sigma_{x,1}\mathrm{d}W_1 + \Sigma_{x,2}\mathrm{d}W_2, \qquad
\mathrm{d}y = \big[L_y(t,x)\,y + f_y(t,x)\big]\mathrm{d}t
  + \Sigma_{y,1}\mathrm{d}W_1 + \Sigma_{y,2}\mathrm{d}W_2 .$$

The coefficients may be arbitrarily nonlinear in $(t, x)$; what is
forbidden is nonlinearity in $y$. Two Wiener channels are carried
separately so that observation and signal noise can be correlated, which
the filter's gain accounts for.

### Writing a model

`aci_model()` takes the blocks directly:

```{r cgns-model}
m2 <- aci_model(
  Lx  = function(t, x) matrix(c(1, 0), 2, 1),
  fx  = function(t, x) -0.5 * x,
  Ly  = function(t, x) matrix(-0.5, 1, 1),
  fy  = function(t, x) 0,
  Sx1 = function(t, x) matrix(c(0.5, 0.3, 0, 0.4), 2, 2),
  Sy2 = function(t, x) matrix(1, 1, 1),
  k = 2, l = 1, name = "correlated observation noise")
m2
```

`aci_model_from_affine()` takes the full drift instead and does the split for
you, after checking that the split exists. A drift that is not affine in
the hidden state is rejected rather than silently linearised:

```{r affine-reject, error = TRUE}
aci_model_from_affine(
  f_full    = function(t, x, y) -0.5 * x + y^2,
  g_full    = function(t, x, y) -0.5 * y,
  Sx        = function(t, x) matrix(0.5, 1, 1),
  Sy_hidden = function(t, x) matrix(1, 1, 1),
  k = 1, l = 1)
```

```{r affine-accept}
hand_dyad <- aci_model_from_affine(
  f_full    = function(t, x, y) -0.5 * x + 2 * x * y + 0.5,
  g_full    = function(t, x, y) -0.5 * y - 2 * x^2 + 1,
  Sx        = function(t, x) matrix(0.5, 1, 1),
  Sy_hidden = function(t, x) matrix(1, 1, 1),
  k = 1, l = 1, name = "hand-built dyad")
hand_dyad
```

`aci_linear_model()` is the shortcut for the linear-Gaussian case,
where the two coupling blocks and the two diffusions are constants rather
than functions of $(t, x)$. It expands scalars to the right shapes and returns
an ordinary `cgns_model` object, so the engine takes it like any other:

```{r linear-shortcut}
lin <- aci_linear_model(lambda_x = 1, lambda_y = -0.5,
                        fx = 0, fy = 0,
                        sigma_x = 0.5, sigma_y = 1,
                        k = 1, l = 1)
lin
c(Lx = lin$Lx(0, 1), Ly = lin$Ly(0, 1))

s_lin <- simulate(lin, seed = 2, T = 4, dt = 0.01, burn_in = 1)
max(aci(lin, as_obs(s_lin), init = list(mean = 0, cov = matrix(1, 1, 1)))$aci)
```

Coefficient functions are **mathematical coefficients**: for a fixed
$(t,x)$ they must return deterministic values with stable shapes. The
closed-form routes may realise each coefficient once on the observation
grid and reuse that realised path, so random draws or result-changing
mutable state inside a coefficient function are outside the contract.

### Benchmark constructors

Three constructors carry the reference MATLAB codebase's models:
`aci_dyad_model()`, `aci_enso_model()` and `aci_predprey_model()`. Each records
its
provenance and how far the correspondence has been checked;
`vignette("vignette-3-matlab")` goes through those records model by model,
and exercises the predator-prey partitions.

```{r constructors}
sapply(list(dyad          = aci_dyad_model(),
            enso6         = aci_enso_model(hidden = "tau"),
            predator_prey = aci_predprey_model(hidden = "prey")),
       function(mm) c(observed = mm$k, hidden = mm$l))
```

Two of those records matter for the rest of this vignette. First,
`aci_enso_model()` covers five hidden partitions, and the fifth, $T_C$ hidden,
is
an approximation rather than an exact split: it is built only when the
substitution is named, it assimilates a reduced two-channel observed process,
and it refuses to simulate. Second, **coefficient agreement and inference
agreement are one claim; pathwise simulator agreement is a separate one, and
only the first is made here.** `simulate()` is Euler-Maruyama throughout, and
`meta$matlab_simulator_parity` records where the reference does something
else.

### Observations

`observed_trajectory()` derives `dt` from the supplied times and refuses a
non-uniform grid. A desynchronised time column is an error, not a wrong
answer:

```{r obs-contract, error = TRUE}
observed_trajectory(c(0, 0.01, 0.03), matrix(c(1, 2, 3), ncol = 1))
```

Naming the columns is what makes the conditional questions of section 4
readable, because `aci_conditional()` accepts channel names.


## 2. The engine: filter, smoother, metric

```{r engine-setup}
m    <- aci_dyad_model()
sim  <- simulate(m, seed = 7, T = 8, dt = 0.01, burn_in = 1)
ob   <- observed_trajectory(sim$obs$t, sim$obs$x, names = "x")
init <- list(mean = 0, cov = matrix(1, 1, 1))

filt <- aci_filter(m, ob, init = init)
smoo <- aci_smoother(m, ob, filter = filt)
a    <- aci(m, ob, init = init)
```

### Covariances are strict

A covariance that leaves the positive-definite cone inside a state
recursion, a metric input or the likelihood **stops the run**. The condition
is classed and names the site, the grid index, the time and the offending
value, rather than reporting a number the recursion did not produce. Take
the same record on a twentyfold coarser grid, where the explicit Riccati
step overshoots:

```{r strict-cov, error = TRUE}
idx    <- seq.int(1L, length(ob$t), by = 20L)
ob_bad <- observed_trajectory(ob$t[idx], ob$x[idx, , drop = FALSE],
                              names = "x")
aci_filter(m, ob_bad, init = init)
```

`regularize = "floor"` is the previous behaviour, available on `aci()`,
`aci_filter()`, `aci_smoother()` and `lag_table()`, and
`options(aci.regularize = "floor")` sets it session-wide. It does not make
the run correct; it makes what it did visible:

```{r regularization-floor}
bad <- aci_filter(m, ob_bad, init = init, regularize = "floor")
bad$meta$regularization[c("policy", "fired", "n_events")]
bad$meta$regularization$sites
```

The record is always present, so a clean run says so explicitly rather than
by omission:

```{r regularization-record}
a$meta$regularization[c("policy", "fired", "n_events")]
nrow(a$meta$regularization$sites)
```

`aci_range()` on an `aci_result` reads the policy off the result rather
than off the option, so re-analysing a saved result in a later session
cannot silently change it. `spd_floor()` and `safe_chol()` are exported and
unchanged; they are the implementation of the opt-in, not a default. Called
directly, `spd_floor()` lifts eigenvalues to a floor and `safe_chol()` factors
through a jitter ladder, and gives up rather than returning a factor of
something that was never a covariance:

```{r spd-utilities}
spd_floor(matrix(c(1, 0, 0, -1e-14), 2, 2))
safe_chol(matrix(c(2, 1, 1, 2), 2, 2))
try(safe_chol(matrix(-1, 1, 1)))
```

### Steppers

The default explicit stepper is the reference scheme, and it is the
convention the lag table of section 3 assumes. A positivity-preserving
implicit stepper is available for stiff models or very diffuse priors,
with `nsub` sub-steps per observation:

```{r stepper}
filt_i <- aci_filter(m, ob, init = init, stepper = "implicit", nsub = 5)
a_i    <- aci(m, ob, init = init, stepper = "implicit", nsub = 5)

c(filter_mean_gap = max(abs(filt$mean - filt_i$mean)),
  aci_gap         = max(abs(a$aci - a_i$aci)),
  peak_aci        = max(a$aci))
```

The two steppers are different discretisations, so they give different
numbers, and the difference here is not negligible against the size of the
metric itself. Pick one deliberately and record which.

`lag_table()` refuses anything else outright, because the Theorem 3
recursions it accumulates are exact for the explicit single-step
discretisation and for no other:

```{r stepper-reject, error = TRUE}
lag_table(m, ob, mode = "forward", init = init, stepper = "implicit")
```

### The predictive likelihood

The filter chain can accumulate a predictive log-likelihood. ACI never
uses it, so it is switchable:

```{r loglik}
filt$meta$loglik
is.null(aci_filter(m, ob, init = init, loglik = FALSE)$meta$loglik)
```

`loglik = FALSE` leaves every ACI quantity unchanged and skips the work.
Adoption ledger entry C1b records the measured saving on the warm filter.

### Supplying a filter to the smoother

`aci_filter()` seals the path it returns with a private token, and
`aci_smoother()` skips the per-step covariance re-validation of a path that
still authenticates against the run being smoothed. Any other supplied
path, including one that has been through `saveRDS()`, is validated in
full as before, and `force_validate = TRUE` validates unconditionally.
The smoother result is the same either way:

```{r trusted-filter}
identical(aci_smoother(m, ob, filter = filt)$mean,
          aci_smoother(m, ob, filter = filt, force_validate = TRUE)$mean)
```

Ledger entry C2a records why this was worth doing: on a 3001-point dyad,
validation was 97.7% of `aci_smoother()`'s time when the documented
`aci_filter()` then `aci_smoother()` pattern was used.

### The metric, and its parts

`aci_metric_pair()` and `aci_metric()` are exposed directly, so the
metric can be recomputed from posteriors obtained any other way:

```{r kl}
aci_metric_pair(mu_p = 0, R_p = matrix(1, 1, 1),
                mu_q = 0.5, R_q = matrix(2, 1, 1))
max(abs(aci_metric(smoo, filt)$total - a$aci))
```

The signal and dispersion parts are each floored at zero independently
before being summed, so a decomposition never reports a negative part from
round-off.


## 3. The forward causal influence range

### Two routes, and what they cost

The finite-lag divergences can be stored or streamed.

* `lag_table()` **retains** the rows. Cost is proportional to the retained
  lag cells, which is $O(N^2)$ for a full table, and the object holds
  them. The table can then be reduced repeatedly at no further cost.
* `aci_range()` on an `aci_result` **streams**, forming and reducing one
  row at a time and keeping none. Memory is linear.

Both reduce identically. On this record, asking the streamed route for a
subsample of anchors costs a fraction of the whole-record computation:

```{r cir-routes}
anchors <- seq(1L, length(ob$t), by = 8L)
fc <- aci_range(a, anchors = anchors)
l1 <- aci_range(a, anchors = anchors, method = "l1_linf")
fc
```

`anchors` selects **which anchor times are reported**, and the streamed
engine forms only those rows, skipping the per-interval primitives before
the earliest of them. Asking for fifteen anchors instead of every one is a
saving in work, not a subsample taken afterwards, and the values it returns
are the same ones to the bit:

```{r cir-window}
ob_w <- observed_trajectory(ob$t[1:401], ob$x[1:401, , drop = FALSE],
                            names = "x")
a_w  <- aci(m, ob_w, init = init)
w15  <- seq(1L, 401L, by = 28L)

t_all <- system.time(f_all <- aci_range(a_w))[["elapsed"]]
t_w15 <- system.time(f_w15 <- aci_range(a_w, anchors = w15))[["elapsed"]]

c(all_401_anchors = t_all, fifteen_anchors = t_w15, ratio = t_all / t_w15)
identical(f_all$tau[w15], f_w15$tau)    # the same values, not a resample
```

The saving is on the streamed route only. `lag_table()` still builds every
row, so a retained table plus `aci_range(anchors = )` pays the whole
quadratic cost and saves only the reduction; adoption ledger entry C2b
records that split and the measurements behind it.

What `anchors` does **not** do is cap how far forward each row looks. A
per-anchor lookahead cap is a separate argument, discussed at the end of
this section, and it is not in this build.

### Table storage arguments

When the table is retained, three arguments govern how much of it is kept.
They trade storage and time against the tail of each row:

* `tol` is the level below which a row becomes a candidate for freezing.
* `window` is how many consecutive steps a row must stay below `tol`
  before it is actually frozen, which stops a single dip from truncating a
  row that later recovers.
* `max_lag` is a hard cap on the retained positive lag.

Truncation is not free and it is not hidden. Measured against an
untruncated reference on a shorter record:

```{r table-storage}
ob_s <- observed_trajectory(ob$t[1:301], ob$x[1:301, , drop = FALSE],
                            names = "x")
reduce_tab <- function(...) {
  tb <- lag_table(m, ob_s, mode = "forward", init = init, ...)
  f  <- aci_range(tb, min_M = 0)
  c(mean_retained_lag = round(mean(tb$L), 1),
    Mb                = round(as.numeric(object.size(tb)) / 2^20, 2),
    max_tail_bound    = signif(max(f$tail_bound), 2),
    tau               = list(f$tau))
}
untruncated <- reduce_tab(tol = 0)
compare <- function(...) {
  r <- reduce_tab(...)
  c(unlist(r[1:3]),
    max_tau_error = signif(max(abs(r$tau - untruncated$tau), na.rm = TRUE), 2))
}
rbind(`tol=0 (reference)` = compare(tol = 0),
      `tol=1e-2`          = compare(tol = 1e-2),
      `tol=1e-1`          = compare(tol = 1e-1),
      `max_lag=50`        = compare(tol = 0, max_lag = 50))
```

Three accessors read a retained table without reaching into its internals.
`lt_diag()` returns the zero-lag diagonal, `lt_row()` returns one anchor's row
padded out to full length, and `lt_tail_bound()` returns the heuristic tail
estimate. `lt_row(pad = "na")` is the one to reach for when the question is
*where the table stopped* rather than what the row was worth, because the
zero padding of the default is indistinguishable from a genuinely decayed
tail:

```{r table-accessors}
tb_s <- lag_table(m, ob_s, mode = "forward", init = init, tol = 1e-2)
r10  <- lt_row(tb_s, 10, pad = "na")

c(row_cells_padded_out = length(r10),
  cells_actually_kept  = sum(!is.na(r10)),
  retained_lag_L       = tb_s$L[10])
lt_tail_bound(tb_s, 10)
round(utils::head(lt_row(tb_s, 10), 5), 4)
```

Truncation is recorded on the reduced result as well: `bound` says whether
the number came off a truncated table, and `tail_bound` estimates what the
frozen tails would have contributed.

```{r table-bound}
c(bound = fc$bound, l1_bound = l1$bound)
max(fc$tail_bound)
```

`tail_bound` is an **estimate, not a certified error bound**. In the table
above it tracks the realised error to within an order of magnitude for the
`tol` rule, and overstates it for the hard `max_lag` cap. Setting
`options(aci.default_tol = 0)` disables truncation and reproduces the
untruncated convention of the published figures. Note that the streamed
route reads `tol` from that option rather than from an argument.

`mode = "full"` retains the negative-lag cells as well, at
correspondingly more work, and requires `max_lag = Inf`.

### Three named functionals

The objective range can be asked for in three ways, and the names matter
more than usual here.

* `method = "exact"` (the default) is the **definitional objective**: the
  subjective range averaged over every threshold. Read off the running
  maximum of the row, that average collapses to a finite sum,
  `dt * sum(suffix max) / M`. There is no quadrature in it, which is why
  `quadrature = "simpson"` and `quadrature = "sum"` return the same
  number for this method.
* `method = "l1_linf"` is the **efficient ratio** the reference script
  computes, `dt * integral(row) / M`.
* `method = "exact", quadrature = "matlab_eps_grid"` is the
  **MATLAB-compatibility mode**, reproducing the reference script's
  `defn_objective_CIR` by taking a Simpson quadrature of the subjective
  read-out over a threshold grid.

The first two are different functionals, not two quadratures of one; they
coincide only where the divergence decreases with lag. On this record they
differ by a factor of

```{r functional-gap}
round(max(fc$tau / l1$tau, na.rm = TRUE), 3)
```

This is the naming hazard the comparison work flagged explicitly
(`reports/C2_dyad-online-cir.md`, "Naming hazards"): the word "objective"
in the predecessor R implementation names the *ratio*, not the
definitional form. Mapping by name alone pairs quantities that differ by a
factor of up to 1.9 on the shared fixture. Read `method`, `bound` and
`meta$quadrature` off the result rather than trusting a remembered
default.

The first and third are the *same* functional on different axes. The
compatibility mode carries a quadrature error that the exact sum does not,
and refining its grid moves it towards the exact value:

```{r matlab-mode}
mg   <- aci_range(a, anchors = anchors, quadrature = "matlab_eps_grid")
mg4k <- aci_range(a, anchors = anchors, quadrature = "matlab_eps_grid",
                  epsilon_grid = 10^seq(-6, 0.5, length.out = 4097))
mg$meta$quadrature
c(default_513_nodes = max(abs(mg$tau   - fc$tau), na.rm = TRUE),
  refined_4097      = max(abs(mg4k$tau - fc$tau), na.rm = TRUE))
```

Use it for parity with the reference script; use the default for anything
else.

Two arguments fix the ratio's quadrature:

* `quadrature`: `"simpson"` (the default) follows the ACI reference code's
  active line, `"sum"` is the literal L1 grid-function sum used by the
  FBCIR scripts.
* `simpson_close`: the closing rule for the leftover interval when a grid
  has an even number of points. `"quadratic"` reproduces `simps.m`,
  `"trapezoid"` reproduces results reported by earlier package versions.

```{r quadrature}
sum_rule <- aci_range(a, anchors = anchors, method = "l1_linf",
                      quadrature = "sum")
c(max_abs = max(abs(sum_rule$tau - l1$tau), na.rm = TRUE),
  max_rel = max(abs(sum_rule$tau - l1$tau) / l1$tau, na.rm = TRUE))
```

A quadrature choice is worth tens of percent here, so it is not a
formatting detail. `simpson_close` only bites on rows with an even number
of points, so on a record whose rows are all odd-length the two closing
rules agree exactly; that is a property of the record, not evidence that
the argument is inert.

### The subjective range, and the two read-out conventions

The $\varepsilon$-resolved **subjective** range, the lag of the last
exceedance of each threshold, is reached through `epsilon`:

```{r subjective}
sc <- aci_range(a, anchors = anchors,
                epsilon = 10^seq(-4, 0, length.out = 30))
dim(sc$subjective)
sc$meta$convention
```

There are two published read-outs, one grid step apart, and `convention`
names them:

* `"count"` (the default) is the ACI reference script's
  `subj_CIR_idx * dt`, the 1-based index of the last exceedance times the
  step.
* `"lag_time"` is the paper convention, the lag time of that exceedance
  with the first cell at lag 0.

```{r convention}
lag_time <- aci_range(a, anchors = anchors,
                      epsilon = 10^seq(-4, 0, length.out = 30),
                      convention = "lag_time")
gap <- sc$subjective - lag_time$subjective
c(dt              = sc$dt,
  min_gap         = min(gap),
  max_gap         = max(gap),
  only_0_or_dt    = all(abs(gap) < 1e-12 | abs(gap - sc$dt) < 1e-12))
```

The difference is exactly `dt`, or zero where nothing exceeded the
threshold. Neither is an error; the one-step offset was verified to the
unit in last place across 51471 cells in `reports/C1_enso-joint.md`.

**`epsilon` and `epsilon_grid` are separate arguments on purpose.**
`epsilon` reports thresholds; `epsilon_grid` supplies quadrature nodes to
the compatibility mode. Asking one vector to do both jobs drove the
predecessor implementation's objective **negative** on a sparse grid
(`reports/C2_dyad-online-cir.md`, hazard 3), so the package refuses the
combination outright:

```{r epsilon-roles, error = TRUE}
aci_range(a, anchors = anchors, epsilon_grid = c(1e-3, 1e-2))
```

### Statuses and masking

```{r statuses}
table(fc$status)
sum(is.na(fc$tau))
```

`status` classifies each anchor, and the four levels are not four grades of
the same thing: two of them report a range and two of them decline to.

* `resolved`: the record comfortably outlasts the influence it measured, so
  the reported range is supported by the observations that follow the anchor.
* `censored`: it does not. The last exceedance of the strength floor sits in
  the later half of the row, leaving less record after the influence than the
  influence itself occupied, and the reported range should be read as a lower
  bound. The test is scale-free and taken entirely from the row; no horizon or
  margin is supplied by the caller.
* `below_threshold`: the row's peak divergence never cleared the floor, so the
  range is masked rather than reported as an inflated number.
* `insufficient`: fewer than three observations follow the anchor. The final
  anchor is always `insufficient`.

Two of those bear expanding.

**Censoring is judged inside the row, not against a caller-supplied
horizon.** The obvious test, "did the exceedance run into the last cell",
can never fire: the last cell of a complete row is identically zero,
because given the whole record the online estimate is the smoother it is
scored against. The test used instead compares the record remaining after
the last exceedance against the extent of the exceedance itself. A record
that does not outlast the influence it measured did not resolve it, and
its range is a lower bound. The decline in $\tau_f$ at the end of a record
is the same finite-record artefact seen from the other side.

**Masking is not smoothing.** `min_M` masks anchors whose peak divergence
is too small for the ratio to mean anything, `masked_value = "na"` keeps
that visible, and `masked_value = "zero"` restores the published zero
convention. A masked anchor is a refusal to report, not an estimate of
zero influence.

### Not in this build: a per-anchor lookahead cap

`anchors` caps how many rows are formed. It does not cap how far forward
each of them looks, and on a long record the far tail of a row contributes
nothing but cost. On the streamed route the only present control over row
length is the truncation rule, through `options(aci.default_tol = )`; on a
retained table it is `max_lag`, which applies to the whole table rather
than per anchor. A lookahead cap of the second kind is specified in the
0.1.0 plan and is not implemented here.

Anyone adding it should note that `lag_table(window = )` already means
something else entirely, the consecutive-steps-below-`tol` freeze rule of
the previous section, and that the reporting selection is already called
`anchors`. A third meaning for either name would be a trap.


## 4. Conditional ACI

### The construction

With several observed effects, "does $y$ influence the observations"
becomes "does $y$ influence $x_A$ once $x_B$ is accounted for".
`aci_conditional()` names $x_B$, the **complement** of the causal question, and
offers two constructions:

* `method = "mask"` keeps $x_B$ in the dynamics but gives its
  innovations no assimilation weight;
* `method = "reduce"` substitutes $x_B$ as known forcing and
  removes it from the observed block entirely.

These are different operators. `method = "reduce"` is admissible only
when the observation-noise Gram has no $A$-$B$ cross-block, and the
package checks that along the whole path rather than at its first point:

```{r cross-noise, error = TRUE}
s2 <- simulate(m2, seed = 1, T = 1, dt = 0.01)
o2 <- observed_trajectory(s2$obs$t, s2$obs$x, names = c("xA", "xB"))
aci_conditional_reduce(m2, o2, aci_conditional("xB", "reduce"))
```

`method = "mask"` carries no such restriction and works on the same model:

```{r cross-noise-mask}
init2 <- list(mean = 0, cov = matrix(1, 1, 1))
c(total = max(aci(m2, o2, init = init2)$aci),
  conditional = max(aci(m2, o2, init = init2,
                        conditional = aci_conditional("xB", "mask"))$aci))
```

On models whose observation noise *is* diagonal and uncorrelated with the
hidden noise, the two methods agree. That was verified bit-for-bit across
the ENSO partitions (`reports/C3_scalar-partitions.md`,
`reports/C4_conditional.md`) on the shared comparison record. On the shorter
$h_W$ record built here, which the rest of this section also uses, the two
paths agree to round-off rather than bitwise: the arithmetic takes a
different route to the same operator, not to a second answer.

```{r strategies-coincide}
me     <- aci_enso_model(hidden = "hW")
se     <- simulate(me, seed = 12, T = 4, dt = 5e-3, burn_in = 1)
obe    <- as_obs(se)
init_e <- list(mean = me$meta$ic_default$y0, cov = matrix(0.1, 1, 1))

a_inf <- aci(me, obe, init = init_e,
             conditional = aci_conditional(target = "TC", method = "mask"))
a_pf  <- aci(me, obe, init = init_e,
             conditional = aci_conditional(target = "TC",
                                           method = "reduce"))
max(abs(a_inf$aci - a_pf$aci))
```

That agreement is a property of this model's noise structure, not a general
identity, and it should not be carried over to a model whose noise structure
has not been checked. On `m2` above, whose observation noise has an $A$-$B$
cross-block, `method = "reduce"` is not merely different; it is refused.

### State the estimand

Each of the reference conditional scripts fixes a specific target. The
$h_W$ script studies $h_W \to T_C \mid (u, T_E, \tau, I)$, with two
further targets present but commented out. `aci_conditional()` takes either side
of that split, `target` or `given`, never both, and the constructor records
which estimand the script asked for and where it read it:

```{r estimand}
me$meta$causal_link
me$meta$estimand_provenance

by_target     <- aci_conditional(target = "TC", method = "mask")
by_complement <- aci_conditional(given = me$meta$vars$observed[
                             me$meta$conditioning_obs_idx],
                           method = "mask")
by_target
by_complement

a_total <- aci(me, obe, init = init_e)
a_tc    <- aci(me, obe, init = init_e, conditional = by_target)
a_comp  <- aci(me, obe, init = init_e, conditional = by_complement)

c(total = max(a_total$aci), target_TC = max(a_tc$aci))
identical(a_tc$aci, a_comp$aci)         # one split, named from either side
```

The choice is not cosmetic. Conditioning removes a real part of the metric
here, but only because of *which* channel it removes. The hidden state
reaches the observed drift only through the non-zero rows of $L_x$:

```{r estimand-Lx}
Lx_max <- sapply(seq_along(obe$t),
                 function(j) abs(drop(me$Lx(obe$t[j], obe$x[j, ]))))
setNames(round(apply(Lx_max, 1, max), 4), me$meta$vars$observed)
```

Three of the five rows are identically zero, so conditioning on those three
is not a small correction, it is bit-for-bit the unconditional run:

```{r estimand-inert}
a_inert <- aci(me, obe, init = init_e,
               conditional = aci_conditional(given = c("u", "tau", "I"),
                                             method = "mask"))
identical(a_inert$aci, a_total$aci)     # masking {u, tau, I} changes nothing
max(abs(a_total$aci - a_tc$aci))        # removing TE changes this much
```

Whether a conditional mask bites is a property of the model's coupling
structure, and it is worth checking on any new model before reading a
conditional result as evidence. Note also what the metadata is and is not:
`causal_link` and `conditioning_obs_idx` record the estimand the source
script defines, and nothing is conditioned on until a specification is
supplied to `aci()` or declared by the constructor.

### Two conditional conventions covered elsewhere

Two further conditional choices are convention-matching questions rather than
API questions, and `vignette("vignette-3-matlab")` treats them with the rest of
the reference-codebase material: `aci_conditional(first_step = "matlab")`, which
reproduces the reference scripts' unmasked first slice of the
observation-precision array, and `aci_enso_model(observations = )`, which
selects
between the $\tau$-hidden script's three-channel estimand and the full
five-channel construction. Both change the numbers reported, and the second
changes the estimand rather than the estimator.


## 5. Two smoothers, two schemes

The package carries two smoothing **discretisations**, and they are not
interchangeable at finite $\Delta t$.

* `aci_smoother()` is an Euler discretisation of the **continuous** backward
  equation. This is what `aci()` uses for its headline metric.
* The **discrete** Theorem 3 posterior is what `lag_table()` accumulates and
  what `aci_online()` returns, and it is what every CIR quantity is scored
  against.

They agree only to first order in $\Delta t$. Every object records which
discretisation produced it, in `meta$scheme`, and an `aci_result` reads that
into `meta$smoother_scheme`. The finer implementation tag is separate, in a
path's `meta$route`, so the label a reader needs and the label a maintainer
needs are not the same field:

```{r schemes-labels}
tb <- lag_table(m, ob, mode = "forward", init = init, max_lag = 1)

c(aci          = a$meta$smoother_scheme,
  aci_smoother    = smoo$meta$scheme,
  lag_table    = tb$meta$scheme,
  aci_online    = aci_online(m, ob, lag = Inf, init = init)$meta$scheme)
c(smoother_route = smoo$meta$route,
  table_reference_smoother = tb$meta$reference_smoother)
```

`max_lag = 1` is the cheap way to obtain the discrete diagonal without
building a table: it retains only the cells needed for the zero-lag entry,
so the cost is linear in the record rather than quadratic, and the
diagonal is the same one a full table would give.

The gap between the two schemes on this record:

```{r schemes-gap}
c(max_abs_gap = max(abs(lt_diag(tb) - a$aci)),
  peak_aci    = max(a$aci))
```

This is a real difference between two discretisations of the same object,
not a defect in either. The comparison work measured it on the ENSO record
in both R implementations independently and got the **same number to all
printed digits**, 1.04276e-01 on a scale of 1.09324
(`reports/C1_enso-joint.md`, section 1), which is what establishes it as a
scheme property rather than a package difference.

Feeding a table to `aci()` swaps the metric onto the discrete scheme
deliberately, and the result says so:

```{r schemes-swap}
a_discrete <- aci(m, ob, init = init, table = tb)
c(engine = a_discrete$meta$engine,
  scheme = a_discrete$meta$smoother_scheme)
```

When comparing an ACI curve against a published figure, check which
smoother the figure used before concluding anything from a discrepancy of
this size.

### The fixed-lag online smoother

`aci_online()` is the discrete scheme exposed on its own, at any lag. The
estimate at index $j$ conditions on the observed record through
$j + \mathtt{lag}$ and saturates at the end of the record, which the result
records per anchor:

```{r online-surface}
on <- aci_online(m, ob, lag = 20, init = init)
on
on$meta$lag
c(saturated_anchors = sum(on$meta$saturated),
  lag_at_the_start  = on$meta$lag_effective[1],
  lag_at_the_end    = utils::tail(on$meta$lag_effective, 1))
```

The requested lag is not always the lag obtained: the last anchors have no
record left to condition on, so `lag_effective` falls away to zero and
`saturated` marks where it did. Nothing is padded or extrapolated to hide
that.

The two ends are exact, and they are exact against different things.
`lag = 0` returns the filter moments value for value, and `lag = Inf`
returns the complete Theorem 3 posterior, which is the lag table's own
reference smoother rather than `aci_smoother()`:

```{r online-ends}
on_full <- aci_online(m, ob, lag = Inf, init = init)

identical(aci_online(m, ob, lag = 0, init = init)$mean, filt$mean)
c(full_lag_vs_table_diagonal =
    max(abs(aci_metric(on_full, filt)$total - lt_diag(tb))),
  full_lag_vs_da_smooth = max(abs(on_full$mean - smoo$mean)))
```

That last number is the scheme gap again, seen on the means instead of on
the metric. **`aci_online(lag = Inf)` does not reproduce `aci_smoother()`, and
it is not meant to.** The gap grows with the length of the record rather
than staying a fixed offset: on the packaged ENSO partition the smoothed
means differ by up to 1.89e-02 over 401 steps and 9.58e-02 over 4001, and
the ACI values by 0.104 against a scale of 1.093 and 0.482 against 2.347
(adoption ledger entry C3f).

`lag` has no default, deliberately: the lag is the argument the function
exists for, and a defaulted full-lag result is exactly the one that would be
mistaken for `aci_smoother()`. Cost is linear in the record whatever the lag,
so choosing the lag is a modelling decision and not a budget:

```{r online-cost}
lags <- c(1, 20, 400, 800)
setNames(sapply(lags, function(L)
  system.time(aci_online(m, ob, lag = L, init = init))[["elapsed"]]),
  paste0("lag_", lags))
```

An online path carries `kind = "online"`, and that is what keeps it out of
the places a complete smoother is required: `lag_table(smoother = )` rejects
it rather than treating a truncated lag as the whole record.

```{r online-kind, error = TRUE}
lag_table(m, ob, mode = "forward", init = init,
          filter = filt, smoother = on)
```


## Evidence

The measurements cited above by name are in the comparison reports
`C1_enso-joint.md`, `C2_dyad-online-cir.md`, `C3_scalar-partitions.md` and
`C4_conditional.md`; the adopted implementation changes are recorded, with
their accuracy gates, in the package's adoption ledger. Numbers computed
in this vignette are computed on this vignette's own short records and are
not the same measurements.

Those reports and fixtures are not all the same kind of evidence, and the
differences matter for what any of them establishes.
`vignette("vignette-3-matlab")` sets out the grades.
