---
title: "Reproducing the reference MATLAB codebase"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml.js"
    pandoc_args: ["--include-in-header=mathjax-config.html"]
vignette: >
  %\VignetteIndexEntry{Reproducing the reference MATLAB codebase}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 3.6, collapse = TRUE,
                      comment = "#>")
library(acir)
```

*Assimilative causal inference (ACI) in R* (`vignette("vignette-1-intro")`)
runs the engine and *The closed-form ACI engine*
(`vignette("vignette-2-advanced")`) documents its arguments. This one is about
the codebase underneath both: the reference MATLAB implementation `ACI_code`,
which models and which numbers this package reproduces from it, where it
deliberately departs, and how far each correspondence has actually been
checked.

Two claims are kept apart throughout. **Coefficient and inference agreement**
says that given the same driving path, the package's realised model
coefficients, filter, smoother and metric match the script's. **Pathwise
simulator agreement** says that `simulate()` generates the same trajectories
the script does. Only the first is claimed here, and only where a constructor
says so.


## 1. What is claimed, model by model

Three constructors carry the reference codebase's models. Each records its
provenance and how far the correspondence has been checked:

```{r provenance}
aci_dyad_model()$meta$provenance
aci_dyad_model()$meta$source_status
aci_enso_model(hidden = "tau")$meta$source_partition
aci_enso_model(hidden = "tau")$meta$matlab_simulator_parity
aci_predprey_model(hidden = "prey")$meta$provenance
aci_predprey_model(hidden = "prey")$meta$source_status
aci_predprey_model(hidden = "prey")$meta$matlab_simulator_parity
```

`matlab_simulator_parity = FALSE` on the ENSO family is a deliberate, recorded
limitation rather than an omission: `simulate()` is Euler-Maruyama throughout,
while the MATLAB ENSO scripts mix Euler updates for the interannual variables
with Milstein-style updates for $I$ and $\tau$.

### The predator-prey partitions

The stochastic Lotka-Volterra example of the supplementary material is
supplied as two constructors rather than one, because the MATLAB file contains
sequential direction-specific blocks: one run treats the prey as hidden, a
later one treats the predator as hidden, and they are separate causal
questions on the same system. The package keeps them separate, and the metric
answers each on its own model:

```{r predator-prey}
mp_prey <- aci_predprey_model(hidden = "prey")
mp_pred <- aci_predprey_model(hidden = "predator")
c(prey_hidden = mp_prey$meta$vars$hidden,
  prey_observed = mp_prey$meta$vars$observed,
  predator_hidden = mp_pred$meta$vars$hidden,
  predator_observed = mp_pred$meta$vars$observed)

sp_prey <- simulate(mp_prey, seed = 11, t_end = 20, dt = 0.01, burn_in = 5)
sp_pred <- simulate(mp_pred, seed = 11, t_end = 20, dt = 0.01, burn_in = 5)
a_prey  <- aci(mp_prey, as_obs(sp_prey),
               init = list(mean = mp_prey$meta$ic_default$y0,
                           cov = matrix(1, 1, 1)))
a_pred  <- aci(mp_pred, as_obs(sp_pred),
               init = list(mean = mp_pred$meta$ic_default$y0,
                           cov = matrix(1, 1, 1)))
c(peak_prey_hidden = max(a_prey$aci), peak_predator_hidden = max(a_pred$aci))
```

The two numbers are not comparable to each other. They are peaks of two
different metrics, on two different hidden states, along two different paths.
Reading one against the other would be reading a direction off an accident of
scaling. Compare each against its own decoupled behaviour instead.

Note what the metadata carries here: `aci_predprey_model()` records
`source_status = "paper + MATLAB checked"` (its coefficients are graded
against the authors' source in the oracle suite), and
`matlab_simulator_parity = TRUE` records that `simulate()` and
`noisy_predator_prey_model.m` use the same Euler-Maruyama scheme. The
`simulation_convention` field states the limit of that claim: scheme
agreement only, since the RNG streams differ, and no pathwise parity is
claimed.


## 2. The one partition that is an approximation, and says so

`aci_enso_model()` covers the five hidden partitions of the reference codebase.
Four of them split the six-state drift exactly. The fifth, $T_C$ hidden,
cannot: the damping $c_1(t, T_C)\,T_C$ is cubic in $T_C$, so the system is
not conditionally Gaussian with $T_C$ withheld, and the reference script
restores conditional linearity by freezing $c_1$ at the climatology
$T_C = 0$. The constructor requires that substitution to be named, and the
model it returns assimilates only $(T_E, I)$, with $u$, $h_W$ and $\tau$
supplied as prescribed forcings looked up by index on their own grid:

```{r tc-partition}
mj   <- aci_enso_model(hidden = c("u", "hW", "tau"))
sj   <- simulate(mj, seed = 21, t_end = 4, dt = 5e-3, burn_in = 1)
path <- data.frame(t = sj$obs$t, u = sj$hidden[, 1], hW = sj$hidden[, 2],
                   tau = sj$hidden[, 3])

mtc <- aci_enso_model(hidden = "TC", approximation = "zeroth_order_c1",
                      prescribed = path)
mtc
c(observed = paste(mtc$meta$vars$observed, collapse = ", "),
  prescribed = paste(mtc$meta$vars$prescribed, collapse = ", "),
  approximation = mtc$meta$approximation)
```

The refusal is shown through a small helper that catches the classed
condition and prints its class and message, so the reader sees what the
package says rather than a halted chunk:

```{r refused}
refused <- function(expr) {
  tryCatch({
    force(expr)
    cat("no condition was signalled\n")
  }, error = function(e) {
    cat(class(e)[1L], ": ", conditionMessage(e), "\n", sep = "")
  })
}
```

What the filter and smoother return are the moments of that *approximating*
model, not of the six-state system, so the model refuses to generate a path
of its own and names the constructor that can:

```{r tc-simulate}
refused(simulate(mtc, seed = 1, t_end = 1, dt = 5e-3))
```


## 3. `matlab_defect_compat`, and the defect it names

`matlab_defect_compat = TRUE` reproduces one further departure: the script's
assimilation forcing omits a thermocline term its own simulator drift
carries. The omission moves the means and leaves both covariance sequences
untouched, because the term enters only the mean equations:

```{r tc-defect}
obc    <- observed_trajectory(sj$obs$t,
                              cbind(TE = sj$obs$x[, "TE"], I = sj$obs$x[, "I"]))
init_c <- list(mean = sj$obs$x[1, "TC"], cov = matrix(0.1, 1, 1))
mtd    <- aci_enso_model(hidden = "TC", approximation = "zeroth_order_c1",
                         prescribed = path, matlab_defect_compat = TRUE)

f_inc  <- aci_filter(mtc, obc, init = init_c)   # term included
f_omit <- aci_filter(mtd, obc, init = init_c)   # script verbatim
a_inc  <- aci(mtc, obc, init = init_c)
a_omit <- aci(mtd, obc, init = init_c)

identical(f_inc$cov, f_omit$cov)        # an identity, not a tolerance
c(filter_mean_gap  = max(abs(f_inc$mean - f_omit$mean)),
  aci_gap          = max(abs(a_inc$aci - a_omit$aci)),
  integrated_ratio = sum(a_omit$aci) / sum(a_inc$aci))
```

The default is the corrected term, not the script. That is a deliberate
choice and it is the one place in the package where the default is *not* the
reference behaviour: the term is available, its omission is inconsistent with
the same script's own simulator, and the sibling scripts carry it. Anyone
grading against published `T_C` output should set `matlab_defect_compat =
TRUE` and say so.


## 4. The first-slice convention in the conditional scripts

The reference conditional scripts fill the first slice of the
observation-precision array with the **full** Gram inverse and mask only the
later slices. That is a convention, not a kernel failure, and `first_step`
names both sides of it. `"uniform"` masks every slice and is the default, and
`"matlab"` reproduces the asymmetry:

```{r first-step}
me     <- aci_enso_model(hidden = "hW")
se     <- simulate(me, seed = 12, t_end = 4, dt = 5e-3, burn_in = 1)
obe    <- as_obs(se)
init_e <- list(mean = me$meta$ic_default$y0, cov = matrix(0.1, 1, 1))

by_target <- aci_conditional(target = "TC", method = "mask")
by_matlab <- aci_conditional(target = "TC", method = "mask",
                             first_step = "matlab")

f_u  <- aci_filter(me, obe, init = init_e, conditional = by_target)
f_m  <- aci_filter(me, obe, init = init_e, conditional = by_matlab)
a_tc <- aci(me, obe, init = init_e, conditional = by_target)
a_m  <- aci(me, obe, init = init_e, conditional = by_matlab)

c(step2_filter_mean_gap = abs(f_u$mean[2] - f_m$mean[2]),
  step2_relative        = abs(f_u$mean[2] - f_m$mean[2]) / abs(f_u$mean[2]),
  last_step_gap         = abs(f_u$mean[nrow(f_u$mean)] -
                              f_m$mean[nrow(f_m$mean)]),
  peak_aci_gap          = max(a_tc$aci) - max(a_m$aci),
  integrated_relative   = sum(a_m$aci) / sum(a_tc$aci) - 1)
```

One slice out of `r length(obe$t)` moves the run by that much, and the
difference decays through the record rather than vanishing: it is still
above round-off at the last step. `"matlab"` applies to
`method = "mask"` only, which is where a masked precision path exists,
and it is exactly inert wherever the mask itself is.


## 5. The $\tau$ partition: two estimands, not two implementations

The $\tau$-hidden reference script prescribes $u$ and $h_W$ as known
series and runs a three-channel observed process, stating that this
"will yield the same results" as the five-channel form. On our path it
does not. Both prescribed channels have non-zero $L_x$ rows, so they
carry genuine $\tau$ information, and prescribing them reproduces their
effect on the drift but not their innovations.

That is why the estimand is a constructor argument here rather than a
caller's option. `observations = "reduced"` is the script's estimand and the
default for this partition, carried as a specification the assimilation runs
whenever the caller supplies none, and `"full"` is the five-channel
construction:

```{r tau-estimands}
mt_red  <- aci_enso_model(hidden = "tau")
mt_full <- aci_enso_model(hidden = "tau", observations = "full")
c(default = mt_red$meta$observations, alternative = mt_full$meta$observations)
mt_red$meta$estimand_nontarget

st     <- simulate(mt_full, seed = 3, t_end = 8, dt = 5e-3, burn_in = 1)
obt    <- as_obs(st)
init_t <- list(mean = mt_red$meta$ic_default$y0, cov = matrix(0.1, 1, 1))

a_reduced <- aci(mt_red,  obt, init = init_t)
a_full    <- aci(mt_full, obt, init = init_t)

Lx_tau <- sapply(seq_along(obt$t),
                 function(j) abs(drop(mt_full$Lx(obt$t[j], obt$x[j, ]))))
setNames(round(apply(Lx_tau, 1, max), 5), mt_full$meta$vars$observed)

c(peak_full        = max(a_full$aci),
  peak_reduced     = max(a_reduced$aci),
  max_abs_diff     = max(abs(a_full$aci - a_reduced$aci)),
  mean_ratio       = mean(a_reduced$aci) / mean(a_full$aci),
  pearson          = cor(a_full$aci, a_reduced$aci))
```

Observations are supplied on all five channels either way. What changes is
which of them are assimilated and which are read as known forcing.

The independent measurement on the shared comparison record, kept in the
development record (section 7), found the same shape at
larger amplitude: time-resolved ACI differing by up to 0.776, which is
3.1 times its own mean level, with Pearson correlation 0.905, while the
*time-averaged* ACI agreed to within 0.5%. The shortcut roughly preserves
the average level while materially distorting the time-resolved curve,
and the time-resolved curve is what the method reports.

The two forms are therefore **different estimands**, and any fidelity
claim has to say which observation set it reproduces. `meta$observations`
records which one a model carries, and the development record has the
measurements behind the choice of default.


## 6. Reading the conventions off a result

Reproducing a published number means matching every convention that number
was computed under, and the package's position is that those conventions
belong on the result object rather than in a reader's memory. The fields to
check, and where each is explained:

```{r conventions-checklist}
m    <- aci_dyad_model()
sim  <- simulate(m, seed = 7, t_end = 4, dt = 0.01, burn_in = 1)
ob   <- as_obs(sim)
init <- list(mean = 0, cov = matrix(1, 1, 1))
a    <- aci(m, ob, init = init)
fc   <- aci_range(a, anchors = seq(1L, length(ob$t), by = 20L))

c(smoother_scheme = a$meta$smoother_scheme,
  regularization  = a$meta$regularization$policy,
  cir_method      = fc$method,
  cir_bound       = fc$bound,
  cir_quadrature  = fc$meta$quadrature,
  cir_convention  = fc$meta$convention)
```

* `smoother_scheme` distinguishes the continuous backward equation from the
  discrete Theorem 3 posterior. They differ at first order in $\Delta t$ and
  the difference is not small. *The closed-form ACI engine*, section 5,
  measures it.
* `regularization` is `"none"` on a strict run and `"floor"` where the caller
  opted into projecting covariances back into the positive-definite cone. On a
  `"floor"` run the same list records how many times it fired and where.
* `method` and `bound` distinguish the definitional objective from the
  reference script's efficient ratio, and say whether the table behind the
  number was truncated. The two functionals are not interchangeable, and the
  word "objective" names the ratio in the predecessor R implementation.
* `quadrature` covers the `"matlab_eps_grid"` compatibility mode and the
  `"simpson"` / `"sum"` split between the ACI and FBCIR reference scripts.
* `convention` distinguishes the reference script's index-count read-out of
  the subjective range from the paper's lag time, which differ by exactly one
  grid step.

The arguments behind the CIR-side fields are documented in *The closed-form
ACI engine*, section 3, with the size of each change measured there. What this section adds is only the instruction to read them
off the object before comparing anything against a figure.


## 7. What the evidence is, and what it is not

Fidelity claims in this package are graded by the kind of evidence behind
them, and the kinds are not interchangeable.

* An **authors-source fixture** is numerical output produced by the reference
  MATLAB itself. It is the strongest evidence available and the package has
  it only where the source distribution shipped it.
* A **source-derived harness** drives the package with inputs read out of the
  reference scripts and compares against quantities the scripts define. It
  tests the translation, not the original execution.
* An **independent transcription** re-implements a script's numerics from the
  source text in R, separately from the package, and compares the two. The
  $T_C$ partition's fixtures are of this kind: their oracle is the script
  `make-tc-fixtures.R` under the repository's `tools/fixtures/`, and no
  MATLAB was executed to make them. Stronger than package-to-package agreement, weaker than an
  authors-source fixture.
* **Analytic identities** are checks that must hold by construction, such as
  the two covariance sequences of section 3 being bit-identical because the
  omitted term enters only the mean equations. These are asserted as
  identities rather than as tolerances.
* **Package-to-package agreement** between two R implementations sharing a
  fixture set is shared evidence, not a second oracle. It is useful for
  locating a disagreement and it establishes nothing about MATLAB fidelity on
  its own.
* **Behavioural tests** fix the package's own contracts, such as which inputs
  are refused. They say nothing about the reference at all.

The measurements cited above by name are in the repository's development
record, `dev/acir-process-rationale.md`, which records each adopted change
with the measurement and the accuracy gate behind it. Numbers computed in
this vignette are computed on its own short records and are not those
measurements.
