---
title: "What the future tells you about the present"
subtitle: "A walkthrough of the aciR package, from one function call to the point where it refuses to answer"
date: 2026-08-17
date-format: iso
abstract: |
  Assimilative causal inference measures how much the later part of an
  observed signal sharpens an estimate of something that was never observed.
  This walkthrough climbs five rungs, from a single function call on a
  two-variable system to a six-dimensional climate model, and ends at the
  boundary where the package reports that the record ran out before the
  question could be settled. Every number and figure below is produced by
  running the code shown, on systems that are published and cited.
format:
  html:
    embed-resources: true
    # MathML rather than the default MathJax. MathJax is fetched from a content
    # delivery network at view time, which a self-contained page must not do,
    # and omitting it silently leaves the LaTeX on the page as literal text.
    # MathML is rendered by the browser itself, so the page stays offline and
    # the notation still resolves. The key is `html-math-method`, which is what
    # Quarto passes to pandoc; `math_method` is the rmarkdown spelling and is
    # accepted silently here without doing anything.
    html-math-method: mathml
    toc: true
    toc-location: left
    toc-depth: 2
    number-sections: true
    code-fold: false
    code-tools: false
    fig-width: 8
    fig-height: 4.2
    fig-align: center
    theme:
      light: cosmo
    smooth-scroll: true
    include-in-header:
      text: |
        <style>
          :root {
            --rung:  #1f4e79;
            --warn:  #b45309;
            --good:  #146b52;
            --paper: #fbfcfd;
          }
          body { font-size: 1.02rem; }
          h1, h2, h3 { letter-spacing: -0.01em; }
          h2 { margin-top: 2.4rem; padding-bottom: .35rem;
               border-bottom: 2px solid #e6ebf1; }
          .lede { font-size: 1.12rem; line-height: 1.62; color: #35414d;
                  margin: 0 0 1.4rem 0; }
          .box { border-left: 4px solid var(--rung); padding: .75em 1.1em;
                 margin: 1.35em 0; background: var(--paper);
                 border-radius: 0 4px 4px 0; }
          .box-warn { border-left-color: var(--warn); background: #fffaf3; }
          .box-good { border-left-color: var(--good); background: #f3faf7; }
          .box p:last-child, .box ul:last-child { margin-bottom: 0; }
          .box .hd { font-weight: 700; display: block; margin-bottom: .3em;
                     color: var(--rung); letter-spacing: .02em;
                     text-transform: uppercase; font-size: .78rem; }
          .box-warn .hd { color: var(--warn); }
          .box-good .hd { color: var(--good); }
          .rung-tag { display: inline-block; font-size: .72rem; font-weight: 700;
                      letter-spacing: .09em; text-transform: uppercase;
                      color: #6b7885; border: 1px solid #d4dbe3;
                      border-radius: 999px; padding: .16em .8em;
                      margin-bottom: .5rem; }
          details > summary { cursor: pointer; font-weight: 600;
                              padding: .35em 0; }
          table.dataTable { font-size: .9em; }
          pre { border-radius: 4px; }
          .quiet { color: #667381; font-size: .93rem; }
        </style>
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    message: false
    warning: false
    dev: "png"
    dpi: 120
    fig.retina: 1
---

```{r setup, include=FALSE}
library(aciR)
library(ggplot2)
library(DT)

set.seed(2026)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "#eaeef2"),
      plot.title = element_text(face = "bold", size = 12.5),
      plot.subtitle = element_text(colour = "#5b6672", size = 10.5),
      axis.title = element_text(colour = "#42505e"),
      legend.position = "top",
      legend.title = element_blank()
    )
)

# Okabe-Ito, chosen because it stays legible under every common form of
# colour vision deficiency.
pal <- c("#0072B2", "#D55E00", "#009E73", "#CC79A7")

# Every fixture read below ships inside the installed package, so this
# document renders from a clean R session with nothing downloaded.
ex <- function(f) system.file("extdata", f, package = "aciR")

tbl <- function(df, n = nrow(df), digits = 4) {
  num <- vapply(df, is.numeric, logical(1L))
  df[num] <- lapply(df[num], function(z) signif(z, digits))
  DT::datatable(
    utils::head(df, n),
    rownames = FALSE,
    options = list(pageLength = min(n, 12L), dom = "t", scrollX = TRUE),
    class = "compact stripe hover"
  )
}
```

::: lede
Some things you want to know about cannot be measured. You watch one signal
and you want the state of another that no instrument reaches. The question
this package answers is narrow and answerable: given everything the observed
signal does *later*, how much better can you say what the unobserved state
was *now*, and how far ahead do you have to look before the answer stops
improving?
:::

## What this document is

Assimilative causal inference is a method for conditional Gaussian nonlinear
systems, published by Andreou et al. (2026). A conditional Gaussian nonlinear
system is one in which an observed signal evolves nonlinearly, while the
unobserved component, *conditional on the observed path*, has Gaussian
statistics available in closed form. That closed form is what makes the
method exact rather than approximate.

Two quantities follow from it. The **causal-information metric** is the
relative entropy, also called the Kullback-Leibler divergence, of the
smoother posterior from the filter posterior at each instant. In words: how
much the future of the observed record sharpens the estimate. The **causal
influence range** is the second quantity, and it asks how far ahead one must
look before that sharpening stops.

::: {.box .box-warn}
[On the inputs]{.hd}
The systems below are published mathematical models, and the paths are
simulated realisations of them or fixtures generated from the reference
implementation. **They are not observational records.** In particular, the
El Niño Southern Oscillation section uses a simulated realisation of a
published climate model, not measured sea-surface temperatures, and nothing
here supports an empirical claim about the climate. The models, and the
findings that use them, are in the source paper.
:::

::: {.box}
[Requirements]{.hd}
`aciR` (0.2.3 or later), `ggplot2` and `DT`. Nothing is downloaded: every
fixture read below ships inside the installed package. To install:

```r
# install.packages("remotes")
remotes::install_github("biometryhub/ACI", subdir = "aciR")
```
:::

## Rung 1: the whole method in one call {#rung-1}

[Rung 1 of 5]{.rung-tag}

The nonlinear dyad is the smallest system worth the trouble. One observed
variable, one unobserved variable, and a coupling that produces intermittent
bursts rather than steady behaviour. Those bursts are the interesting part,
because that is where the future of the observed signal has the most to say.

`aci()` runs the whole workflow: the forward filter, the backward smoother,
and the metric that contrasts them.

```{r rung1-run}
model <- aci_dyad_model()
sim <- aci_simulate(model, n = 6000, seed = 42)
result <- aci(sim$x, model, dt = 0.001)

result
```

The `print` method leads with the distribution of the metric rather than a
single number, because a single number would hide the thing that matters:
the metric is small most of the time and large in short episodes.

```{r rung1-fig, fig.height=4.6}
d <- as.data.frame(result)

p1 <- ggplot(d, aes(t, x)) +
  geom_line(linewidth = .32, colour = pal[1]) +
  labs(title = "The observed signal", y = "x", x = NULL)

p2 <- ggplot(d, aes(t, aci)) +
  geom_area(fill = pal[2], alpha = .22) +
  geom_line(linewidth = .38, colour = pal[2]) +
  labs(
    title = "What the future adds",
    subtitle = "Causal-information metric, in nats",
    y = "metric", x = "time"
  )

patch <- function(a, b) {
  gridExtra::grid.arrange(a, b, ncol = 1, heights = c(1, 1.15))
}
if (requireNamespace("gridExtra", quietly = TRUE)) {
  patch(p1, p2)
} else {
  print(p2)
}
```

The metric peaks where the signal does something abrupt. That is the
substantive reading: during a burst, the record that arrives afterwards
carries a great deal of information about what the unobserved component was
doing at the time, and during quiet stretches it carries much less.

## Rung 2: how far ahead must you look? {#rung-2}

[Rung 2 of 5]{.rung-tag}

The metric says *how much*. It does not say *how long*. Two systems can
sharpen an estimate by the same total amount, one settling within a few
steps and the other still improving thousands of steps later. The causal
influence range separates them.

```{r rung2-run}
comp <- aci_dyad_components(sim$x, model$parameters)
filt <- aci_filter(sim$x, comp, dt = 0.001, mu0 = model$y0, R0 = 0.1)

cir <- aci_cir(
  sim$x, comp,
  dt = 0.001, filt = filt,
  window = seq(500L, 3000L, by = 250L)
)

cir
```

Two objective ranges are reported, and they differ here by more than a
factor of two. This is not a numerical disagreement. They are different
functionals that coincide only under a condition this system does not meet.

```{r rung2-tbl}
cd <- as.data.frame(cir)
tbl(cd[, c("time", "objective", "objective_exact", "peak",
           "status", "monotone")])
```

::: {.box}
[Why the two ranges differ]{.hd}
`objective_exact` averages the per-threshold ranges over the whole threshold
grid, which is the definition in the source paper. `objective` is the
efficient approximation the paper also gives. They agree when the divergence
sequence decreases with lag, by the layer-cake identity. The `monotone`
column reports that condition, and here it is `FALSE` at every reported
time, so the gap is definitional rather than a defect in either. The package
reports both instead of picking one, because which is wanted depends on
whether you need the definition or the cheaper computation.
:::

## Rung 3: the question is not symmetric {#rung-3}

[Rung 3 of 5]{.rung-tag}

Nothing so far distinguishes "what x says about y" from "what y says about
x". A stochastic predator-prey system makes the asymmetry concrete, because
the same pair of populations can be read in either causal direction. Observe
the prey and infer the predator, or observe the predator and infer the prey.

```{r rung3-run}
pp <- read.csv(ex("predprey_signal.csv"))
params <- aci_predprey_model()$parameters

run_direction <- function(direction) {
  observed <- if (direction == "predator_to_prey") pp$prey else pp$predator
  cmp <- aci_predprey_components(observed, params, direction)
  f <- aci_filter(observed, cmp, dt = 0.005, mu0 = 4, R0 = 0.1)
  s <- aci_smoother(observed, cmp, dt = 0.005, filt = f)
  data.frame(
    direction = direction,
    observed = if (direction == "predator_to_prey") "prey" else "predator",
    mean = mean(aci_metric(f, s)),
    peak = max(aci_metric(f, s)),
    drift_low = min(cmp$L_y),
    drift_high = max(cmp$L_y)
  )
}

both <- do.call(rbind, lapply(
  c("predator_to_prey", "prey_to_predator"), run_direction
))
tbl(both)
```

The mean metric differs by roughly a factor of two between the two readings
of one system. Watching the predator tells you more about the prey than the
reverse, at these parameters.

This system also exercises something the dyad does not. The unobserved
component's own drift coefficient is set by the observed state, so it moves
over the record and changes sign, shown by the `drift_low` and `drift_high`
columns. The latent process is damped at some times and driven at others,
which a system with a constant coefficient cannot express.

## Rung 4: six dimensions, and what one channel contributes {#rung-4}

[Rung 4 of 5]{.rung-tag}

The El Niño Southern Oscillation model in the source paper is the largest
system the package expresses: six dimensions, three observed and three
unobserved, with coefficients that vary seasonally and noise covariances
that vary in time.

```{r rung4-setup}
enso <- read.csv(ex("enso_signal.csv"))
p <- aci_enso_parameters()
emodel <- aci_enso_model(p)

c(observed = paste(emodel$observed, collapse = ", "),
  unobserved = paste(emodel$unobserved, collapse = ", "))
```

The observed channels are the central- and eastern-Pacific temperature
anomalies and the Walker circulation strength. The unobserved ones are the
zonal current, the thermocline depth and the wind burst.

```{r rung4-joint}
ecomp <- aci_enso_components(
  enso$T_C, enso$T_E, enso$I, p = p, time = enso$t, dt = 0.005
)
xm <- rbind(enso$T_C, enso$T_E, enso$I)
mu0 <- c(6.9136e-04, -0.0028, -0.0256)
R0 <- 0.01 * diag(3L)

ef <- aci_filter(xm, ecomp, dt = 0.005, mu0 = mu0, R0 = R0)
es <- aci_smoother(xm, ecomp, dt = 0.005, filt = ef)
joint <- aci_metric(ef, es)

c(mean = mean(joint), peak = max(joint), at_time = enso$t[which.max(joint)])
```

::: {.box .box-warn}
[Six dimensions defeat the eye]{.hd}
At three observed and three unobserved dimensions there is no plot that
settles whether this is right. A time series of the metric can be shown, and
is below, but it is orientation rather than proof. What establishes the
numbers is the parity grading described at the end of this document, in
which each quantity is compared against the authors' own code on the same
input. Treat the figure as a picture of the answer, not as evidence for it.
:::

`aci_conditional()` asks a sharper question: what does one observed channel
contribute, when the others are still present in the system but the filter
stops assimilating them?

```{r rung4-conditional}
channels <- c("T_C", "T_E", "I")

contribution <- do.call(rbind, lapply(seq_along(channels), function(i) {
  cc <- aci_conditional(ecomp, target = i)
  f <- aci_filter(xm, cc, dt = 0.005, mu0 = mu0, R0 = R0)
  s <- aci_smoother(xm, cc, dt = 0.005, filt = f)
  m <- aci_metric(f, s)
  data.frame(channel = channels[i], mean = mean(m), peak = max(m))
}))
contribution <- rbind(
  contribution,
  data.frame(channel = "all three", mean = mean(joint), peak = max(joint))
)
tbl(contribution)
```

```{r rung4-fig, fig.height=3.9}
ggplot(
  data.frame(t = enso$t, metric = joint),
  aes(t, metric)
) +
  geom_line(linewidth = .34, colour = pal[3]) +
  labs(
    title = "Causal-information metric, six-dimensional model",
    subtitle = "Orientation only. The grading, not the picture, is the evidence.",
    x = "time", y = "metric (nats)"
  )
```

::: {.box .box-warn}
[The trap in that table]{.hd}
The central-Pacific channel on its own scores *higher* than all three
channels together. That is not a contradiction, and it is the single easiest
result to misread. The metric measures how much the future improves on the
filter. Give the filter fewer channels and it becomes worse, which leaves
the future more room to improve on it. A large value therefore reports a
weak filter as readily as an informative channel, and the quantity is not a
decomposition: the parts do not sum to the whole.

The Walker circulation scoring near zero should be read with the same care.
It says the *future* of that channel adds little beyond its own past, not
that the channel is uninformative.
:::

The estimand is worth stating exactly, because four different quantities get
confused here. The metric is the relative entropy of
$p(y_t \mid x_{0:T})$ from $p(y_t \mid x_{0:t})$. It is not an
interventional quantity, not a Granger test and not transfer entropy,
although all four can be large at the same instants. Conditioning changes
what the filter assimilates, not the system that produced the path.

## Rung 5: where the package refuses {#rung-5}

[Rung 5 of 5]{.rung-tag}

Every method has a boundary. For this one the binding constraint is not
dimension and not runtime: it is how much record remains *after* the moment
you asked about. To measure how far ahead the influence extends, the record
has to extend that far. Ask about a time near the end and it does not.

The honest behaviour is to say so. Here the same system is interrogated at
thirty times, marching toward the end of the record.

```{r rung5-run}
sim2 <- aci_simulate(model, n = 4000, seed = 7)
comp2 <- aci_dyad_components(sim2$x, model$parameters)
filt2 <- aci_filter(sim2$x, comp2, dt = 0.001, mu0 = model$y0, R0 = 0.1)

edge <- aci_cir(
  sim2$x, comp2,
  dt = 0.001, filt = filt2,
  window = seq(1000L, 3900L, by = 100L)
)

ed <- as.data.frame(edge)
ed$lookahead <- nrow(sim2) - ed$index
table(ed$status)
```

```{r rung5-fig, fig.height=4.1}
ggplot(ed, aes(lookahead, objective, colour = status)) +
  geom_point(size = 2.1) +
  geom_line(aes(group = 1), linewidth = .3, colour = "#9aa5b1") +
  scale_x_reverse() +
  scale_colour_manual(values = c(resolved = pal[3], censored = pal[2])) +
  labs(
    title = "The boundary is the record that remains, not the model",
    subtitle = paste(
      "Reading right to left, each point sits closer to the end of the record."
    ),
    x = "steps of record remaining after the reported time",
    y = "objective range"
  )
```

The status changes from `resolved` to `censored` partway along. The rule is
explicit: a time is censored when the divergence has not settled below
`threshold` until it has consumed more than `1 - margin` of the record
available after it. At the default `margin = 0.1`, that means more than 90%.

What the package does *not* do at that point is return `NA`. A censored time
still yields a statement the record supports, namely that the range is at
least this long, so the value stands and the status records that it is a
lower bound. Discarding it would throw away information exactly at the end
of the record, which is where somebody studying a recent event is looking.
Only a time with fewer than three later observations, where no quadrature is
possible at all, returns `NA`.

::: {.box .box-good}
[What to take from this rung]{.hd}
A range reported near the end of a record is a lower bound, and the `status`
column is how you tell. Two habits follow. Check `status` before quoting a
range. And when you design a record, leave more tail than the influence you
expect to measure, because the tail is the binding constraint.
:::

## How these numbers are checked {#checked}

A reimplementation tested only against fixtures its own author transcribed
demonstrates self-consistency and nothing more. If the author misread an
equation, the fixture encodes the same misreading and passes.

Every numerical claim in this package is therefore graded against a source
that did not also produce the code under test. The authors publish their
work as MATLAB scripts rather than as a callable library, so their
computational passages are lifted into callable functions as byte-exact
slices, never retyped, and a checker fails on a single byte of drift. Both
implementations then run on the same input and are compared quantity by
quantity.

```{r checked-tbl}
tbl(data.frame(
  check = c(
    "Extract outputs reproducing the reference's own workspace",
    "Verbatim reference lines, none retyped",
    "Scalar core against the reference",
    "Predator-prey, both directions, filter through influence range"
  ),
  result = c(
    "156 of 156, difference exactly 0",
    "2,159 lines across 33 extracts",
    "26 of 26 quantities, worst 1.5e-14",
    "26 graded quantities, worst 1.5e-12"
  )
))
```

For scale, an experiment comparing independently developed implementations
of the same algorithms on identical input found agreement degrading from six
significant figures to one (Hatton, 1997).

The full record is in the
[parity ledger](ledgers/parity_ledger.html), and the design history, with
what remains open, is in the
[development ledger](ledgers/development_ledger.html).

## Where to go next {#next}

```{r next-tbl}
tbl(data.frame(
  rung = c("1", "2", "3", "4", "5"),
  question = c(
    "How much does the future add?",
    "How far ahead must I look?",
    "Does the direction matter?",
    "What does one channel contribute?",
    "When should I not trust the answer?"
  ),
  entry_point = c(
    "aci()", "aci_cir()", "aci_predprey_components()",
    "aci_conditional()", "the status column of aci_cir()"
  )
))
```

Three articles ship with the package. *Assimilative causal inference on the
nonlinear dyad model* is the worked tutorial. *Assumptions and
interpretation* states the conditions under which the method applies and
what a peak does and does not support, and it is the right next read after
the two warnings above. *Validation and the independent oracle* records what
each oracle grades and, for each, what it does not.

## References {#references}

Andreou, M., Chen, N. and Bollt, E. (2026). Assimilative causal inference.
*Nature Communications*, **17**, 1854.
[doi:10.1038/s41467-026-68568-0](https://doi.org/10.1038/s41467-026-68568-0)

Andreou, M., Chen, N. and Li, Y. (2026). An adaptive online smoother with
closed-form solutions and information-theoretic lag selection for
conditional Gaussian nonlinear systems. *Journal of Nonlinear Science*,
**36**(4), 71.
[doi:10.1007/s00332-026-10271-x](https://doi.org/10.1007/s00332-026-10271-x)

Hatton, L. (1997). The T experiments: errors in scientific software.
*IEEE Computational Science and Engineering*, **4**(2), 27-38.
[doi:10.1109/99.609829](https://doi.org/10.1109/99.609829)

## Citation and licence {#licence}

Cite the package as `citation("aciR")`. The method is the authors' and
should be cited alongside any use of this software; the reference is the
first entry above.

`aciR` is MIT licensed, Copyright &copy; 2026 Max Moldovan. The reference
implementation, [`marandmath/ACI_code`](https://github.com/marandmath/ACI_code),
is separately MIT licensed, Copyright &copy; 2025 Marios Andreou, and is not
redistributed with this package. This document inherits the MIT licence of
both.

<details>
<summary>Session information</summary>

```{r session, echo=FALSE}
sessionInfo()
```

</details>
